---
title: "Getting started with rcicr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rcicr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4
)
```

`rcicr` implements **reverse correlation image classification**, a psychophysics technique for visualizing mental representations, for example of faces. It works in two stages:

1. **Stimulus generation.** A base image, such as a face photo, is combined with random visual noise. Each trial shows a pair: the "original" (base plus noise) and its "inverted" counterpart (base minus the same noise). On each trial the participant picks whichever of the two looks more like a target category, such as "trustworthy" or "happy". This is a two-image forced-choice (2IFC) task.
2. **Classification image (CI).** After data collection, the noise of every chosen original is added up and the noise of every chosen inverted image is subtracted. The average is the classification image: it shows which visual features drove the participant's choices.

This vignette runs both stages on a tiny synthetic example. For the full method, with several participants, scaling choices, z-maps and informational value, see `vignette("reverse-correlation-walkthrough", package = "rcicr")`. Example datasets and analysis scripts are in [rcicr_examples](https://github.com/rdotsch/rcicr_examples/).

```{r setup}
library(rcicr)
```

## 1. Generate stimuli

`generateStimuli2IFC()` needs a square base image. To keep this vignette self-contained we make a synthetic greyscale one. In a real study you pass the path to your base face photo(s).

```{r generate-base-face}
set.seed(42)
base_face_path <- tempfile(fileext = ".png")
png::writePNG(matrix(runif(64 * 64), 64, 64), base_face_path)
```

Now generate stimuli for a small task: 20 trials and one base image, at a small size so the vignette builds quickly. A real study typically uses `img_size = 512` and several hundred trials (Dotsch & Todorov, 2012).

```{r generate-stimuli, results = "hide"}
stimulus_path <- tempdir()

generateStimuli2IFC(
  base_face_files = list(face = base_face_path),
  n_trials        = 20,
  img_size        = 64,
  stimulus_path   = stimulus_path,
  seed            = 1,
  ncores          = 1,
  save_as_png     = FALSE # set to TRUE to also write stimulus PNGs to stimulus_path
)

rdata_file <- list.files(stimulus_path, pattern = "\\.Rdata$", full.names = TRUE)[1]
```

This writes an `.Rdata` file to `stimulus_path` holding the noise parameters of every trial. **That file is the only link between stimulus generation and CI computation.** Keep it: every analysis function below needs it as its `rdata` argument.

## 2. Collect (or, here, simulate) responses

In a real experiment you now run the 2IFC task and record, per trial, which image each participant chose: `1` for the original, `-1` for the inverted one. This vignette has no participant, so it simulates random responses. Random responses carry no signal, so the resulting classification image shows nothing; never do this in a real analysis.

```{r simulate-responses}
responses <- sample(c(1, -1), 20, replace = TRUE)
```

## 3. Compute the classification image

`generateCI()` looks up the noise parameters of the stimuli that were shown, weights them by the responses, and averages them into one classification image.

```{r compute-ci}
ci <- generateCI(
  stimuli     = 1:20,
  responses   = responses,
  baseimage   = "face",
  rdata       = rdata_file,
  save_as_png = FALSE
)

names(ci)
```

The result holds four pixel matrices:

- `ci$ci` is the raw noise.
- `ci$scaled` is that noise rescaled for display. The default method, `'independent'`, picks the lowest scaling constant that avoids clipping this particular image; `?generateCI` describes the others.
- `ci$base` is the base image.
- `ci$combined` overlays the scaled noise on the base image.

```{r plot-ci}
image(ci$combined, col = gray.colors(256), axes = FALSE, asp = 1)
```

Because the responses were random, this classification image is just noise. With real data, patterns tied to the participants' choices emerge here.

## Next steps

- `batchGenerateCI()` and `batchGenerateCI2IFC()` compute one CI per participant or condition from a data frame. By default they put the whole batch on one scale with `autoscale()`, so the images can be compared by eye.
- `computeInfoVal2IFC()` computes the informational value: a z-score-like measure of how much signal a CI holds, compared with a simulated null distribution.
- `plotZmap()` shows which regions of a CI carry reliable signal.

Each function's help page, such as `?generateCI` or `?batchGenerateCI`, lists its options and has runnable examples.
