---
title: "Recipes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 2.6, fig.height = 2.6)
oldpar <- par(no.readonly = TRUE)

show <- function(m, title, zlim = range(m, na.rm = TRUE)) {
  op <- par(mar = c(0, 0, 1.4, 0))
  # image() takes [x, y] with y increasing upwards, so a matrix indexed
  # [row, col] has to be transposed and flipped to display the right way up.
  image(t(m[nrow(m):1, ]), col = gray.colors(256), axes = FALSE, asp = 1, # nolint: seq_linter.
        main = title, zlim = zlim, useRaster = TRUE)
  par(op)
}
```

Short answers to questions that come up once the basics in `vignette("reverse-correlation-walkthrough", package = "rcicr")` are in place. Each recipe runs at a small size so this vignette builds in seconds; real studies use `img_size = 512` and several hundred trials.

```{r setup}
library(rcicr)
```

## A stimulus set and a simulated participant

The recipes share one stimulus set: a synthetic face (no photo, so no image licence) and 120 trials.

```{r stimuli, results = "hide"}
n <- 64
x <- matrix(seq(-1, 1, length.out = n), n, n, byrow = TRUE)
y <- -matrix(seq(-1, 1, length.out = n), n, n)
face <- exp(-(x^2 / 0.45 + y^2 / 0.75))
face <- face - 0.55 * exp(-(((x + 0.3)^2 + (y - 0.25)^2) / 0.012))
face <- face - 0.55 * exp(-(((x - 0.3)^2 + (y - 0.25)^2) / 0.012))
face <- face - 0.35 * exp(-(x^2 / 0.10 + (y + 0.42)^2 / 0.006))
face <- (face - min(face)) / (max(face) - min(face))
base_face <- tempfile(fileext = ".png")
png::writePNG(face, base_face)

stimulus_path <- tempfile("stimuli")
generateStimuli2IFC(list(face = base_face), n_trials = 120, img_size = n,
                    stimulus_path = stimulus_path, seed = 1, nscales = 3,
                    ncores = 1, save_as_png = FALSE)
rdata_file <- list.files(stimulus_path, pattern = "\\.Rdata$", full.names = TRUE)[1]
```

So that the results show something, a simulated participant responds according to a hidden template. `evidence` is how strongly each trial's noise matches it.

```{r observer}
e <- new.env()
load(rdata_file, envir = e)
set.seed(99)
template <- generateNoiseImage(rnorm(max(e$p$patchIdx)), e$p)
evidence <- vapply(seq_len(120), function(i) {
  base::sum(generateNoiseImage(e$stimuli_params$face[i, ], e$p) * template)
}, numeric(1))
evidence <- evidence / sd(evidence)
set.seed(7)
choice <- ifelse(evidence + rnorm(120) > 0, 1, -1) # 1: original chosen, -1: inverted
```

## Noise-only stimuli

To show noise without a face (for a task about textures, say, or to inspect the noise itself), use a uniform mid-grey base image. A uniform image has no contrast to maximize, so pass `maximize_baseimage_contrast = FALSE`.

```{r noise-only, results = "hide"}
grey <- tempfile(fileext = ".png")
png::writePNG(matrix(0.5, n, n), grey)
noise_path <- tempfile("noise")
generateStimuli2IFC(list(grey = grey), n_trials = 4, img_size = n,
                    stimulus_path = noise_path, seed = 1, nscales = 3, ncores = 1,
                    maximize_baseimage_contrast = FALSE)
```

```{r noise-only-show, fig.show = "hold"}
ori <- png::readPNG(file.path(noise_path, "rcic_grey_1_00001_ori.png"))
inv <- png::readPNG(file.path(noise_path, "rcic_grey_1_00001_inv.png"))
show(ori, "trial 1, original", zlim = c(0, 1))
show(inv, "trial 1, inverted", zlim = c(0, 1))
```

These are ordinary 2IFC stimuli: the stimulus PNGs average the noise with the base, so on a grey base the noise appears at half contrast, and the `.Rdata` file is analysed as usual. For the noise at full contrast, for display rather than for a task, ask for it back as a data frame, one column per trial:

```{r noise-frame, results = "hide"}
noise <- generateStimuli2IFC(list(grey = grey), n_trials = 4, img_size = n, seed = 1,
                             nscales = 3, ncores = 1, return_as_dataframe = TRUE,
                             maximize_baseimage_contrast = FALSE,
                             save_as_png = FALSE, save_rdata = FALSE)
```

```{r noise-frame-show}
# The same scaling the stimuli use, without the base: most noise falls in [-0.3, 0.3].
noise_1 <- pmin(pmax((matrix(noise[[1]], n, n) + 0.3) / 0.6, 0), 1)
show(noise_1, "trial 1 noise, full contrast", zlim = c(0, 1))
```

## Ratings instead of a two-image choice

`generateCI()` weights each trial's noise by its response, so any numeric response works, not only `1`/`-1`. If participants rated each stimulus on a scale, recode the scale so that its midpoint is 0 and its direction is the one you want the classification image to show. Here a 1–4 scale ("not at all" to "very much") is recoded to -2, -1, 1, 2:

```{r ratings}
set.seed(11)
rating <- as.integer(cut(evidence + rnorm(120, 0, 0.7), c(-Inf, -0.7, 0, 0.7, Inf)))
table(rating)

ci_rating <- generateCI(1:120, c(-2, -1, 1, 2)[rating], "face", rdata_file,
                        save_as_png = FALSE)
cor(as.vector(ci_rating$ci), as.vector(template))
```

**Check the direction.** Recoded the wrong way round, the scale gives the anti-classification image, and that looks like a plausible face too. With real data there is no template to compare against: compare instead with a CI whose direction you know, such as a 2IFC CI for the same target, or check that the expected features appear.

Responses must be numeric and complete. A missing rating, a factor, or `TRUE`/`FALSE` stops `generateCI()` with an error naming the trials, rather than returning an image of `NA`s. Drop unanswered trials from `stimuli` and `responses` alike.

## Matching the InfoVal reference to the design

`computeInfoVal2IFC()` compares a classification image with a reference distribution: CIs built from random responses. The reference must match the CI's design. When it does not, the InfoVal is miscalibrated.

### When some trials were dropped

Say 20 trials were dropped because the participant did not respond in time. The CI is built from 100 stimuli; the default reference uses all 120.

```{r dropped}
kept <- setdiff(1:120, seq(3, 120, by = 6))
ci_kept <- generateCI(kept, choice[kept], "face", rdata_file, save_as_png = FALSE)
attr(ci_kept, "trial_design")$stimuli[1:10]
```

`generateCI()` records the stimuli it used, so the matching reference is one argument away. The default call says the two differ:

```{r dropped-infoval, results = "hide", message = TRUE}
default <- computeInfoVal2IFC(ci_kept, rdata_file, iter = 10000)
matched <- computeInfoVal2IFC(ci_kept, rdata_file, iter = 10000,
                              reference_stimuli = attr(ci_kept, "trial_design")$stimuli)
```

```{r dropped-infoval-show}
c(default = default, matched = matched)
```

Fewer stimuli give a noisier CI, so its norm is larger under random responding too: against the full-set reference, the default InfoVal is inflated. Each reference is simulated once and stored in the `.Rdata` file for later calls.

### A masked classification image

A CI computed with a `mask` holds `NA` in the masked pixels. Its InfoVal is computed over the unmasked pixels, against a reference over the same pixels. Here only the eye region is kept (in a mask, `0` is masked):

```{r masked, results = "hide"}
mask <- matrix(0, n, n)
mask[18:30, 10:54] <- 1
ci_eyes <- generateCI(1:120, choice, "face", rdata_file, mask = mask, save_as_png = FALSE)
infoval_eyes <- computeInfoVal2IFC(ci_eyes, rdata_file, iter = 10000)
```

```{r masked-show}
show(ci_eyes$ci, "CI, eye region only")
infoval_eyes
```

An InfoVal over part of the image is not comparable with one over the whole image.

### Several classification images at once

`batchComputeInfoVal2IFC()` scores a list of CIs, such as the result of `batchGenerateCI2IFC()`, and simulates each distinct reference once:

```{r batch, results = "hide"}
set.seed(3)
data <- data.frame(
  participant = rep(c("p1", "p2", "p3"), each = 120),
  stimulus    = rep(1:120, 3),
  response    = c(ifelse(evidence + rnorm(120, 0, 0.5) > 0, 1, -1),
                  ifelse(evidence + rnorm(120, 0, 2) > 0, 1, -1),
                  sample(c(1, -1), 120, replace = TRUE))
)
cis <- batchGenerateCI2IFC(data, "participant", "stimulus", "response", "face",
                           rdata_file, save_as_png = FALSE)
infovals <- batchComputeInfoVal2IFC(cis, rdata_file, iter = 10000)
```

```{r batch-show}
round(infovals, 2)
```

The third participant answered at random, and scores near 0.

### A CI averaged over participants

No reference is defined for a CI that averages several participants (or repeated presentations of the same stimulus): every reference assumes one response per stimulus from one responder. `computeInfoVal2IFC()` returns a number but says it is not calibrated.

```{r averaged, results = "hide", message = TRUE}
ci_group <- generateCI(data$stimulus, data$response, "face", rdata_file,
                       participants = data$participant, save_as_png = FALSE, n_cores = 1)
infoval_group <- computeInfoVal2IFC(ci_group, rdata_file, iter = 10000)
```

```{r averaged-show}
infoval_group
```

Averaging shrinks the CI's norm, so against a single-responder reference the group CI scores below zero, although one of its participants scored 6.3 on their own. Score each participant's own CI instead, as above.

## Reproducing a number from an earlier version

A classification image or InfoVal reported years ago can be reproduced in two ways.

### Run the version that computed it

This reproduces everything the file already stores exactly. A reference the old release has to simulate anew also needs the original session's `RNGkind()`, which files from before `rng_kind` was stored do not record; R's default has not changed. Use the version your analysis script or its session info records. `p$generator_version` in the `.Rdata` file names the version that *generated the stimuli*, which need not be the one that analysed them. Install that release into a library of its own, so your current installation stays as it is, and load it in a fresh R session:

```{r old-version, eval = FALSE}
dir.create("rcicr-1.4.1")
remotes::install_github("rdotsch/rcicr@v1.4.1", lib = "rcicr-1.4.1")
library(rcicr, lib.loc = "rcicr-1.4.1")
```

Older releases import packages this one no longer needs; `install_github()` installs those too.

### Recompute with the current version

Classification images come out the same unless `NEWS.md` lists your situation under "Reproducibility impact", in the releases after the one you used.

Their file names can differ, though, for per-participant images from a GitHub install of 0.4.1 to 1.2.3. In those versions, a direct `generateCI()` call with `participants` and `save_individual_cis = TRUE` could save an image under another participant's name. Rerunning such a version reproduces the wrong names; recomputing with this one gives the right names, so comparing old and new files by name can turn up differences where there are none. The [individual-CI filename advisory](https://rdotsch.github.io/rcicr/articles/rcicr-individual-ci-advisory.html) explains how to tell whether you are affected. CRAN releases never carried the bug.

An InfoVal depends on the reference distribution stored in the `.Rdata` file:

- **A reference stored by 1.4.0 or later is reused as stored**, so its InfoVal does not change. Such a file holds `reference_norms_source` (see `vignette("stored-data", package = "rcicr")`).
- **An older reference is rebuilt** the first time, with a message that the new InfoVal supersedes earlier ones. By default it is computed through the stimulus Gram matrix, which differs from the calculation of 1.5.0 and earlier only by rounding. `reference_method = "images"` is that earlier calculation, unchanged.
- **A reference rebuilt with `reference_method = "images"` matches the old one exactly** when all of these hold:
    - The old version reconstructed the stimuli correctly. That excludes pre-0.3.0 files, pre-1.1.0 files whose missing settings were replaced by defaults, and pre-1.4.0 references for base images with parameters of their own, which scored every base against the first base's reference.
    - The rebuild uses the original `iter`, and the original `response_seed` if there was one.
    - The random draws use the same generator. Files from before `rng_kind` was stored draw under the session's `RNGkind()`; R's default has not changed, so this matters only if the original session set another kind.
- **A reference drawn with a `response_seed` is kept as stored**, whichever method you ask for.

To rebuild a reference with the earlier calculation, store it with `generateReferenceDistribution2IFC()`, passing the original `iter` (the length of the stored norms, if the file still holds them); later `computeInfoVal2IFC()` calls reuse it. Here both methods score the same classification image, each on its own copy of the file:

```{r reproduce, results = "hide", warning = FALSE}
ci_all <- generateCI(1:120, choice, "face", rdata_file, save_as_png = FALSE)

gram_file <- tempfile(fileext = ".Rdata")
file.copy(rdata_file, gram_file)
generateReferenceDistribution2IFC(gram_file, iter = 1000, ncores = 1)
infoval_gram <- computeInfoVal2IFC(ci_all, gram_file, iter = 1000)

images_file <- tempfile(fileext = ".Rdata")
file.copy(rdata_file, images_file)
generateReferenceDistribution2IFC(images_file, iter = 1000, ncores = 1,
                                  reference_method = "images")
infoval_images <- computeInfoVal2IFC(ci_all, images_file, iter = 1000)
```

```{r reproduce-show}
c(gram = infoval_gram, images = infoval_images, difference = infoval_images - infoval_gram)
```

Real references use `iter = 10000`; the smaller one here keeps the slower `"images"` calculation quick.

```{r, include = FALSE}
par(oldpar)
```
