---
title: "What the wrong-base InfoVal reference does to a conclusion"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "")
options(width = 100)
```

When stimuli are generated for several base images with `use_same_parameters = FALSE`, each base
gets its own parameter matrix, but `generateReferenceDistribution2IFC()` reconstructs the stimuli
without forwarding that setting. The null it builds is therefore always the first base image's
([issue #299](https://github.com/rdotsch/rcicr/issues/299)), and every base after the first is
scored against another base's noise.

This document measures what that costs. It exists because the two obvious readings are both wrong:
"the reference is from the wrong base, so the InfoVal is invalid" overstates it, and "it is only
noise, so it does not matter" understates it. The answer is that the wrong reference is a *valid*
null built on a different realization of the same noise process, so the error is unbiased jitter
whose size is set by the number of trials — and that a significance call flips only in a narrow
window around the cut-off.

Knitting this runs the whole measurement, which takes tens of minutes: `r 30 + 20 + 20 + 12` base
image pairs, each needing two stimulus sets and three reference distributions of 10,000 iterations.

```{r constants}
library(rcicr)

img_size <- 64
nscales <- 5
iter <- 10000
response_seed <- 97
cutoff <- 1.96
```

```{r fixtures}
scratch <- tempfile("infoval-impact")
dir.create(scratch)

base_image <- function(name, seed, size) {
  f <- file.path(scratch, sprintf("%s-%d.png", name, size))
  set.seed(seed)
  png::writePNG(matrix(runif(size^2), size, size), f)
  f
}

# capture.output() here and below only keeps progress bars out of the rendered
# document; the package's functions are called exactly as a user would call them.
generate <- function(bases, n_trials, seed, size, same_parameters) {
  path <- tempfile("stim", tmpdir = scratch)
  dir.create(path)
  invisible(capture.output(
    generateStimuli2IFC(bases, n_trials = n_trials, img_size = size, stimulus_path = path,
                        seed = seed, use_same_parameters = same_parameters, nscales = nscales,
                        ncores = 1, save_as_png = FALSE)
  ))
  list.files(path, pattern = "\\.Rdata$", full.names = TRUE)[1]
}
```

The base images are synthetic noise rather than faces. Nothing here depends on their content: the
classification image and the reference norms are built from the noise alone, and the base image is
added only for display.

## The reference distribution ignores which base you score

Two base images, independent parameters, against the same stimulus set containing only the first of
them. If the reference carried any information about which base is being scored, these two runs
would differ.

```{r mechanism}
a <- base_image("a", 11, img_size)
b <- base_image("b", 12, img_size)

two_bases <- generate(list(a = a, b = b), n_trials = 24, seed = 31, size = img_size,
                      same_parameters = FALSE)
first_only <- generate(list(a = a), n_trials = 24, seed = 31, size = img_size,
                       same_parameters = TRUE)

# iter far below the recommended 10,000 here: this compares two runs against each
# other, not against a cut-off, and the warning is the function's own.
demo_reference <- function(rdata) {
  invisible(capture.output(norms <- suppressWarnings(generateReferenceDistribution2IFC(
    rdata, iter = 200, ncores = 1, response_seed = response_seed, save_rdata = FALSE))))
  norms
}

reference_two <- demo_reference(two_bases)
reference_one <- demo_reference(first_only)

saved <- new.env()
load(two_bases, envir = saved)

max(abs(saved$stimuli_params[["a"]] - saved$stimuli_params[["b"]]))
max(abs(reference_two - reference_one))
```

The two bases hold genuinely different parameters, and the reference distribution is bit-identical
to the one for a file that does not contain the second base at all.

## Base 2's own reference

The correction requires the null the second base should have had: its noise built from the
parameters actually saved for it. `generateNoiseImage()` does the reconstruction, and the loop below
is the arithmetic of `generateReferenceDistribution2IFC()`, batched so that the response draws are
consumed in the same order.

```{r glue}
noise_matrix <- function(params, p, size) {
  vapply(seq_len(nrow(params)),
         function(trial) as.vector(generateNoiseImage(params[trial, ], p)),
         numeric(size^2))
}

reference_norms <- function(noise, iter, seed, block = 500L) {
  set.seed(seed)
  n_trials <- ncol(noise)
  out <- numeric(iter)
  done <- 0L
  while (done < iter) {
    k <- min(block, iter - done)
    responses <- matrix(((runif(n_trials * k) > 0.5) * 2) - 1, nrow = n_trials, ncol = k)
    out[(done + 1):(done + k)] <- sqrt(colSums(((noise %*% responses) / n_trials)^2))
    done <- done + k
  }
  out
}
```

That reimplementation is only trustworthy if it reproduces the package's own output, so every
configuration below checks it against `generateReferenceDistribution2IFC()` on the first base before
using it on the second. The same check is what establishes that the *first* base is scored
correctly: the package's reference matches a null rebuilt from base 1's saved parameters.

```{r measure}
measure_pair <- function(n_trials, seed, size, validate) {
  rdata <- generate(list(a = base_image("a", 11, size), b = base_image("b", 12, size)),
                    n_trials = n_trials, seed = seed, size = size, same_parameters = FALSE)
  saved <- new.env()
  load(rdata, envir = saved)

  noise_a <- noise_matrix(saved$stimuli_params[["a"]], saved$p, size)
  noise_b <- noise_matrix(saved$stimuli_params[["b"]], saved$p, size)

  scored_against <- reference_norms(noise_a, iter, response_seed)   # what the bug uses
  own <- reference_norms(noise_b, iter, response_seed)              # what base 2 should have
  # Same base, an independent draw of the simulated responses: the Monte Carlo
  # wobble InfoVal already carries at this iter, as a yardstick for the shift.
  monte_carlo <- reference_norms(noise_a, iter, response_seed + 1L)

  agreement <- NA_real_
  if (validate) {
    invisible(capture.output(package_reference <- generateReferenceDistribution2IFC(
      rdata, iter = iter, ncores = 1, response_seed = response_seed, save_rdata = FALSE)))
    agreement <- max(abs(scored_against - package_reference))
    stopifnot(agreement < 1e-12)
  }

  data.frame(
    n_trials = n_trials, img_size = size, seed = seed, agreement = agreement,
    d0 = (median(own) - median(scored_against)) / mad(scored_against),
    rho = mad(own) / mad(scored_against) - 1,
    mc_shift = (median(monte_carlo) - median(scored_against)) / mad(scored_against),
    mc_rho = mad(monte_carlo) / mad(scored_against) - 1
  )
}

sweep <- function(n_trials, seeds, size = img_size) {
  do.call(rbind, Map(function(seed, validate) measure_pair(n_trials, seed, size, validate),
                     seeds, seq_along(seeds) == 1))
}
```

`d0` is the shift in the reference median, expressed in the units InfoVal is reported in, and `rho`
is the error in its scale. An InfoVal that should have been `z` is reported as `z + d0 + z * rho`,
so `d0` is what a threshold call sees and `rho` matters only for large values.

```{r sweep, results = "hide"}
results <- rbind(
  sweep(770, 1:30),
  sweep(300, 101:120),
  sweep(100, 201:220),
  sweep(300, 301:312, size = 128)
)
```

```{r agreement}
max(results$agreement, na.rm = TRUE)
```

```{r shift-definitions}
# An InfoVal that should read z is reported as z + shift(). The Monte Carlo
# yardstick has to carry the same scale term, or it is not the same quantity.
shift <- function(rows, z) rows$d0 + z * rows$rho
monte_carlo <- function(rows, z) rows$mc_shift + z * rows$mc_rho

centre <- function(x) {
  se <- sd(x) / sqrt(length(x))
  c(mean = mean(x), se = se, ses_from_zero = abs(mean(x)) / se)
}
```

## The error is unbiased

```{r unbiased}
at_default <- subset(results, n_trials == 770)

knitr::kable(round(rbind(
  "reference median only (d0)" = centre(at_default$d0),
  "scale only (rho)" = centre(at_default$rho),
  "full shift at the cut-off" = centre(shift(at_default, cutoff)),
  "full shift at z = 10" = centre(shift(at_default, 10))
), 4))
```

Base 2's parameters are an independent draw from the same generator as base 1's, so the null it
receives is a legitimate one — just built on another realization of the noise. Across
`r nrow(at_default)` base image pairs at the package defaults the full shift at the cut-off averages
`r sprintf("%+.3f", mean(shift(at_default, cutoff)))`, which is
`r sprintf("%.1f", centre(shift(at_default, cutoff))[["ses_from_zero"]])` standard errors from zero
against a scatter `r sprintf("%.0f", sd(shift(at_default, cutoff)) / abs(mean(shift(at_default, cutoff))))`
times the size of that mean. The bug cannot have moved a set of published InfoVals one way; it
scattered them.

The scale term is tested alongside the median shift because a ratio of two MADs need not average to
one even when both are drawn from the same distribution. Here it does not detectably depart from it
(`r sprintf("%+.4f", mean(at_default$rho))`, `r sprintf("%.1f", centre(at_default$rho)[["ses_from_zero"]])`
standard errors from zero), so the conclusion holds at every InfoVal, not only near the cut-off.

## How large the scatter is

```{r shift-table}
summary_row <- function(rows) {
  dz <- shift(rows, cutoff)
  data.frame(
    n_trials = rows$n_trials[1], img_size = rows$img_size[1], pairs = nrow(rows),
    mean_abs_d0 = round(mean(abs(rows$d0)), 4),
    mean_abs_dz = round(mean(abs(dz)), 4),
    q90_abs_dz = round(unname(quantile(abs(dz), 0.9)), 4),
    max_abs_dz = round(max(abs(dz)), 4),
    monte_carlo = round(mean(abs(monte_carlo(rows, cutoff))), 4)
  )
}

by_config <- split(results, list(results$n_trials, results$img_size), drop = TRUE)
knitr::kable(do.call(rbind, lapply(by_config, summary_row)), row.names = FALSE)
```

`monte_carlo` is the same quantity, computed the same way down to the scale term, for a reference
built from the same base image with a different draw of the simulated responses: the wobble any
InfoVal already carries at `iter` = `r formatC(iter, format = "d", big.mark = ",")`. At the default
`r max(results$n_trials)` trials the bug is worth
`r sprintf("%.1f", mean(abs(shift(at_default, cutoff))) / mean(abs(monte_carlo(at_default, cutoff))))`
times that wobble. It is the same kind of error as the one the method already tolerates, and not far
from the same size.

The scale error `rho` barely exceeds its own Monte Carlo control
(`r sprintf("%.4f", mean(abs(at_default$rho)))` against
`r sprintf("%.4f", mean(abs(at_default$mc_rho)))`), so almost all of the damage is the shift in the
median, and the `z * rho` term stays small until the InfoVal itself is large.

```{r large-z}
at_z <- function(z, f) vapply(z, function(zz) mean(abs(f(at_default, zz))), numeric(1))

z <- c(2, 5, 10)
knitr::kable(data.frame(
  z = z,
  mean_abs_dz = round(at_z(z, shift), 4),
  as_share_of_z = round(at_z(z, shift) / z, 4),
  monte_carlo = round(at_z(z, monte_carlo), 4),
  ratio = round(at_z(z, shift) / at_z(z, monte_carlo), 2)
))
```

The margin over the Monte Carlo wobble narrows as the InfoVal grows, because both quantities pick up
the same scale term. At a large InfoVal the bug is barely distinguishable in size from re-running the
reference simulation with different random responses.

## It is set by trials, not by pixels

```{r scaling}
at_64px <- subset(results, img_size == 64)
by_trials <- do.call(rbind, lapply(split(at_64px, at_64px$n_trials), summary_row))
reference <- by_trials[by_trials$n_trials == 100, "mean_abs_d0"]
knitr::kable(data.frame(
  n_trials = by_trials$n_trials,
  observed = by_trials$mean_abs_d0,
  predicted_from_100 = round(reference * sqrt(100 / by_trials$n_trials), 4)
), row.names = FALSE)
```

The shift falls as one over the square root of the number of trials, which is what a reference
median built from an average over trials should do. Image size does not enter: at 300 trials the
shift is `r sprintf("%.4f", summary_row(subset(results, n_trials == 300 & img_size == 128))$mean_abs_d0)`
at 128px against
`r sprintf("%.4f", summary_row(subset(results, n_trials == 300 & img_size == 64))$mean_abs_d0)` at
64px, equal within the sampling error of
`r nrow(subset(results, n_trials == 300 & img_size == 128))` and
`r nrow(subset(results, n_trials == 300 & img_size == 64))` pairs. So the 770-trial row above stands
for a study run at the default 512px, which is far too slow to measure directly.

This also puts the 24-trial reproduction in the issue in proportion. Extrapolating the fitted
scaling to 24 trials gives a typical shift of
`r sprintf("%.2f", reference * sqrt(100 / 24))`, so the −0.067 observed there is one unremarkable
draw, and a real study with hundreds of trials sits several times below it.

## What it takes to reach a different conclusion

A significance call at `z = 1.96` changes only when the correct InfoVal lies within the shift of the
cut-off *and* the shift points across it. Applying the measured shifts to a stand-in population of
true InfoVals:

```{r flip-rate}
flip_rate <- function(dz, true_z) {
  outcomes <- outer(true_z, dz, `+`)
  mean((true_z > cutoff) != (outcomes > cutoff))
}

set.seed(20260906)
spreads <- list("uniform 0-4" = runif(20000, 0, 4),
                "uniform 0-10" = runif(20000, 0, 10),
                "uniform 1.5-2.5" = runif(20000, 1.5, 2.5))
flipped <- vapply(spreads, function(z) flip_rate(shift(at_default, cutoff), z), numeric(1))
knitr::kable(data.frame(true_infoval_spread = names(spreads), flipped = round(flipped, 4)),
             row.names = FALSE)
```

Those spreads are stand-ins, not measurements: what InfoVals look like across real studies is not
something this package records. They bracket the answer rather than settle it —
`r sprintf("%.1f%%", 100 * flipped[["uniform 0-4"]])` of exposed classification images under a broad
spread, and `r sprintf("%.1f%%", 100 * flipped[["uniform 1.5-2.5"]])` even in the worst case where
every InfoVal in the study is crowded against the cut-off.

And a classification image is only exposed if all of these hold:

1. InfoVal was computed at all — a feature from 0.4.0 onward, while CRAN carried 0.3.4.1 until the
   2021 archival.
2. The stimulus set has two or more base images.
3. It was generated with `use_same_parameters = FALSE`, which is not the default and appears in no
   vignette, example, test, or InfoVal configuration of the release gate.
4. The classification image comes from a base image other than the first — `(k-1)/k` of the images
   in such a study.

## What is not affected

`generateCI()` selects `stimuli_params[[baseimage]]`, so the classification images and z-maps
themselves are built from the right noise. The bug moves one number, the InfoVal, and only for base
images after the first.

## Limits

How often anyone sets `use_same_parameters = FALSE` is unknown. There is no telemetry, and no issue
in the tracker has ever mentioned the setting, so condition 3 above is unquantified rather than
measured to be rare — this document bounds the damage per affected classification image, not the
number of affected studies.

The two references compared here share their simulated response draws, which isolates the noise
realization from Monte Carlo error. A fix that draws base 2's responses from its own position in the
stream adds roughly the `monte_carlo` column on top, which changes none of the conclusions above.

The shift measured here is the correction a fix applies, so it is also the size to quote in a
`NEWS.md` "Reproducibility impact" entry: about
`r sprintf("%.2f", mean(abs(shift(at_default, cutoff))))` in z at the defaults, scaling as one over
the square root of the trial count, unbiased, and affecting no base image but the second and later.
