---
title: "What the wrong-base InfoVal reference does to a conclusion"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "")
options(width = 100)
```

When stimuli are generated for several base images with `use_same_parameters = FALSE`, each base
gets its own parameter matrix, but `generateReferenceDistribution2IFC()` reconstructs the stimuli
without forwarding that setting. The null it builds is therefore always the first base image's
([issue #299](https://github.com/rdotsch/rcicr/issues/299)), and every base after the first is
scored against another base's noise.

This document measures what that costs. It exists because the two obvious readings are both wrong:
"the reference is from the wrong base, so the InfoVal is invalid" overstates it, and "it is only
noise, so it does not matter" understates it. The answer is that the wrong reference is a *valid*
null built on a different realization of the same noise process, so the error is scatter, with a
directional part some two orders of magnitude below it. Its larger half shrinks with the number of
trials; what the rest does is measured below and comes out unresolved. A significance call flips
only in a narrow window around the cut-off.

Knitting this runs the whole measurement, so it is a batch job rather than an interactive one:
`r 30 + 20 + 20 + 12 + 8 + 10` base image pairs, each needing two stimulus sets and three reference
distributions of 10,000 iterations, with fifty of them building two further references of ten times
that size. Those tenfold references and the ten pairs at the 512 pixels a real study uses dominate
the arithmetic in roughly equal measure, and everything else together comes to a fraction of either.

```{r constants}
library(rcicr)

img_size <- 64
nscales <- 5
iter <- 10000
response_seed <- 97
cutoff <- 1.96
```

```{r fixtures}
scratch <- tempfile("infoval-impact")
dir.create(scratch)

base_image <- function(name, seed, size) {
  f <- file.path(scratch, sprintf("%s-%d.png", name, size))
  set.seed(seed)
  png::writePNG(matrix(runif(size^2), size, size), f)
  f
}

# capture.output() here and below only keeps progress bars out of the rendered
# document; the package's functions are called exactly as a user would call them.
# ?generateReferenceDistribution2IFC guarantees the stimuli, and so everything
# measured here, do not depend on ncores. mclapply() below rejects more than one
# core on Windows, which knits there serially rather than not at all.
cores <- if (.Platform$OS.type == "windows") 1L else max(1L, parallel::detectCores() - 1L)

generate <- function(bases, n_trials, seed, size, same_parameters) {
  path <- tempfile("stim", tmpdir = scratch)
  dir.create(path)
  invisible(capture.output(
    generateStimuli2IFC(bases, n_trials = n_trials, img_size = size, stimulus_path = path,
                        seed = seed, use_same_parameters = same_parameters, nscales = nscales,
                        ncores = cores, save_as_png = FALSE)
  ))
  list.files(path, pattern = "\\.Rdata$", full.names = TRUE)[1]
}
```

The base images are synthetic noise rather than faces. Nothing here depends on their content: the
classification image and the reference norms are built from the noise alone, and the base image is
added only for display.

## The reference distribution ignores which base you score

Two base images, independent parameters, against the same stimulus set containing only the first of
them. If the reference carried any information about which base is being scored, these two runs
would differ.

```{r mechanism}
a <- base_image("a", 11, img_size)
b <- base_image("b", 12, img_size)

two_bases <- generate(list(a = a, b = b), n_trials = 24, seed = 31, size = img_size,
                      same_parameters = FALSE)
first_only <- generate(list(a = a), n_trials = 24, seed = 31, size = img_size,
                       same_parameters = TRUE)

# iter far below the recommended 10,000 here: this compares two runs against each
# other, not against a cut-off, and the warning is the function's own.
demo_reference <- function(rdata) {
  invisible(capture.output(norms <- suppressWarnings(generateReferenceDistribution2IFC(
    rdata, iter = 200, ncores = 1, response_seed = response_seed, save_rdata = FALSE))))
  norms
}

reference_two <- demo_reference(two_bases)
reference_one <- demo_reference(first_only)

saved <- new.env()
load(two_bases, envir = saved)

max(abs(saved$stimuli_params[["a"]] - saved$stimuli_params[["b"]]))
max(abs(reference_two - reference_one))
```

The two bases hold genuinely different parameters, and the reference distribution is bit-identical
to the one for a file that does not contain the second base at all.

## Base 2's own reference

The correction requires the null the second base should have had: its noise built from the
parameters actually saved for it. `generateNoiseImage()` does the reconstruction, and the loop below
is the arithmetic of `generateReferenceDistribution2IFC()`, batched so that the response draws are
consumed in the same order.

```{r glue}
noise_matrix <- function(params, p, size) {
  matrix(unlist(parallel::mclapply(seq_len(nrow(params)), mc.cores = cores,
                function(trial) as.vector(generateNoiseImage(params[trial, ], p)))),
         nrow = size^2)
}

# A block holds one column per iteration, so its width comes from the image size,
# keeping the intermediate near 200 MB at any resolution.
reference_norms <- function(noise, iter, seed, block = max(50L, as.integer(2.5e7 / nrow(noise)))) {
  set.seed(seed)
  n_trials <- ncol(noise)
  out <- numeric(iter)
  done <- 0L
  while (done < iter) {
    k <- min(block, iter - done)
    responses <- matrix(((runif(n_trials * k) > 0.5) * 2) - 1, nrow = n_trials, ncol = k)
    out[(done + 1):(done + k)] <- sqrt(colSums(((noise %*% responses) / n_trials)^2))
    done <- done + k
  }
  out
}
```

That reimplementation is only trustworthy if it reproduces the package's own output, so every
configuration below checks it against `generateReferenceDistribution2IFC()` on the first base before
using it on the second. The same check is what establishes that the *first* base is scored
correctly: the package's reference matches a null rebuilt from base 1's saved parameters.

```{r measure}
measure_pair <- function(n_trials, seed, size, validate) {
  rdata <- generate(list(a = base_image("a", 11, size), b = base_image("b", 12, size)),
                    n_trials = n_trials, seed = seed, size = size, same_parameters = FALSE)
  saved <- new.env()
  load(rdata, envir = saved)

  noise_a <- noise_matrix(saved$stimuli_params[["a"]], saved$p, size)
  noise_b <- noise_matrix(saved$stimuli_params[["b"]], saved$p, size)

  scored_against <- reference_norms(noise_a, iter, response_seed)   # what the bug uses
  own <- reference_norms(noise_b, iter, response_seed)              # what base 2 should have
  # Same base, an independent draw of the simulated responses: the Monte Carlo
  # wobble InfoVal already carries at this iter, as a yardstick for the shift.
  monte_carlo <- reference_norms(noise_a, iter, response_seed + 1L)
  # A reference ten times the size stands in for the population one, so the error
  # a single published InfoVal carries can be measured rather than inferred from
  # the rerun difference, which carries the sampling error of two references. Only
  # for the configuration the ratio is quoted at: elsewhere it would cost billions
  # of operations per pair and reach no table.
  decompose <- size == 64 && n_trials %in% c(300, 770)
  precise <- if (decompose) reference_norms(noise_a, 10L * iter, response_seed + 2L) else NULL
  # The same for base 2, so that the difference between the two precise
  # references shows the between-base component with most of the MAD estimation
  # error taken out, and what is left of `rho` can be attributed rather than
  # assumed.
  # Same response draws as `precise`: differencing two independently simulated
  # references would put back the estimation error this comparison exists to
  # remove.
  precise_b <- if (decompose) reference_norms(noise_b, 10L * iter, response_seed + 2L) else NULL
  # The between-base comparison, estimated a second time from a different shared
  # draw. Whatever the two bases really differ by is common to both estimates and
  # cancels in their difference, so that difference is the estimator's own error,
  # measured on the estimator rather than on a differently constructed stand-in.
  precise_alt <- if (decompose) reference_norms(noise_a, 10L * iter, response_seed + 5L) else NULL
  precise_b_alt <- if (decompose) reference_norms(noise_b, 10L * iter, response_seed + 5L) else NULL

  agreement <- NA_real_
  if (validate) {
    invisible(capture.output(package_reference <- generateReferenceDistribution2IFC(
      rdata, iter = iter, ncores = 1, response_seed = response_seed, save_rdata = FALSE)))
    agreement <- max(abs(scored_against - package_reference))
    stopifnot(agreement < 1e-12)
  }

  data.frame(
    n_trials = n_trials, img_size = size, seed = seed, agreement = agreement,
    med_own = median(own), mad_own = mad(own),
    med_scored = median(scored_against), mad_scored = mad(scored_against),
    d0 = (median(own) - median(scored_against)) / mad(scored_against),
    rho = mad(own) / mad(scored_against) - 1,
    mc_shift = (median(monte_carlo) - median(scored_against)) / mad(scored_against),
    mc_rho = mad(monte_carlo) / mad(scored_against) - 1,
    # Oriented like d0 and rho above: the reference that defines the InfoVal on
    # top, the one actually used underneath, so the two are comparable at a z
    # meaning the same thing.
    single_shift = if (is.null(precise)) NA_real_ else
      (median(precise) - median(scored_against)) / mad(scored_against),
    single_rho = if (is.null(precise)) NA_real_ else mad(precise) / mad(scored_against) - 1,
    between_d0 = if (is.null(precise_b)) NA_real_ else
      (median(precise_b) - median(precise)) / mad(precise),
    between_rho = if (is.null(precise_b)) NA_real_ else mad(precise_b) / mad(precise) - 1,
    between_d0_alt = if (is.null(precise_alt)) NA_real_ else
      (median(precise_b_alt) - median(precise_alt)) / mad(precise_alt),
    between_rho_alt = if (is.null(precise_alt)) NA_real_ else
      mad(precise_b_alt) / mad(precise_alt) - 1
  )
}

# Validated on the first pair of each configuration, except at 512px, where
# generateReferenceDistribution2IFC() rebuilds a matrix from its stimulus frame on
# every one of the 10,000 iterations (#306). That frame carries 64 times the pixels
# it does at 64px, where the check is already the slowest thing in this document,
# so the cost there is prohibitive. The reimplementation does not vary with
# resolution.
sweep <- function(n_trials, seeds, size = img_size) {
  do.call(rbind, Map(function(seed, validate) measure_pair(n_trials, seed, size, validate),
                     seeds, seq_along(seeds) == 1 & size < 512))
}
```

`d0` is the shift in the reference median, expressed in the units InfoVal is reported in, and `rho`
is the error in its scale. An InfoVal that should have been `z` is reported as `z + d0 + z * rho`,
so both terms reach a threshold call — `rho` multiplied by the threshold — and the scale term grows
in weight as the InfoVal does.

```{r sweep, results = "hide"}
results <- rbind(
  sweep(770, 1:30),
  sweep(300, 101:120),
  sweep(100, 201:220),
  sweep(300, 301:312, size = 128),
  sweep(770, 501:508, size = 128),
  sweep(300, 401:410, size = 512)
)
```

```{r agreement}
max(results$agreement, na.rm = TRUE)
```

```{r shift-definitions}
# An InfoVal that should read z is reported as z + shift(). The Monte Carlo
# yardstick has to carry the same scale term, or it is not the same quantity.
shift <- function(rows, z) rows$d0 + z * rows$rho
monte_carlo <- function(rows, z) rows$mc_shift + z * rows$mc_rho

centre <- function(x) {
  se <- sd(x) / sqrt(length(x))
  c(mean = mean(x), se = se, ses_from_zero = abs(mean(x)) / se)
}

at_64px <- subset(results, img_size == 64)
```

## The error is scatter, with a directional part too small to matter

```{r unbiased}
# 770 trials at 64 pixels. Not the package's default configuration, whose
# 512 pixels are measured separately below and at fewer trials.
at_770 <- subset(results, n_trials == 770 & img_size == 64)

knitr::kable(round(rbind(
  "reference median only (d0)" = centre(at_770$d0),
  "scale only (rho)" = centre(at_770$rho),
  "full shift at the cut-off" = centre(shift(at_770, cutoff)),
  "full shift at z = 10" = centre(shift(at_770, 10))
), 4))
```

Base 2's parameters are an independent draw from the same generator as base 1's, so the null it
receives is a legitimate one — just built on another realization of the noise. Across
`r nrow(at_770)` base image pairs at 770 trials and 64 pixels the full shift at the cut-off averages
`r sprintf("%+.3f", mean(shift(at_770, cutoff)))`, which is
`r sprintf("%.1f", centre(shift(at_770, cutoff))[["ses_from_zero"]])` standard errors from zero
against a scatter `r sprintf("%.0f", sd(shift(at_770, cutoff)) / abs(mean(shift(at_770, cutoff))))`
times the size of that mean.

Neither term is exactly centred, though, and the sign of each is decided by the same structure.
`rho` is the ratio `mad(own) / mad(scored_against)` less one, and for exchangeable positive scales
`B/A + A/B >= 2` forces `E[B/A] >= 1`, strictly whenever the two differ at all. `d0` divides by the
MAD of one member of the pair only, so its mean is `-Cov(median, 1 / mad)` rather than zero: a noise
realization with more energy has both a larger median norm and a wider spread. Both terms therefore
carry a directional component, and the question is its size rather than its existence.

Neither mean can measure it — both sit far below their own standard errors. Averaging a pair's two
labellings can: had the second base image been listed first, the bug would have scored the first
against *its* null, and the two orderings share the bulk of the variation, which cancels.

```{r directional}
directional <- function(rows, z) {
  d0 <- ((rows$med_own - rows$med_scored) / rows$mad_scored +
         (rows$med_scored - rows$med_own) / rows$mad_own) / 2
  ratio <- 1 + rows$rho
  d0 + z * ((ratio + 1 / ratio) / 2 - 1)
}

knitr::kable(signif(do.call(rbind, lapply(split(at_64px, at_64px$n_trials), function(rows)
  c("median term" = centre(directional(rows, 0))[c("mean", "se")],
    "at the cut-off" = centre(directional(rows, cutoff))[c("mean", "se")],
    "at z = 10" = centre(directional(rows, 10))[c("mean", "se")]))), 3))
```

The estimates come out positive wherever they resolve against their own standard error, which is
what the argument above requires of the expectations, and all of them are minute. At 770 trials and
64 pixels the directional part of the shift is
`r sprintf("%.1e", mean(directional(at_770, cutoff)))` at the cut-off and
`r sprintf("%.3f", mean(directional(at_770, 10)))` at an InfoVal of 10 — around
`r sprintf("%.0f%%", 100 * mean(directional(at_770, 10)) / mean(abs(shift(at_770, 10))))`
of the scatter at that value, a proportion that does not grow with the InfoVal because both scale
with it. A set of published InfoVals was therefore scattered rather than moved, to a precision far
finer than anyone reads these numbers to.

## How large the scatter is

```{r shift-table}
summary_row <- function(rows) {
  dz <- shift(rows, cutoff)
  data.frame(
    n_trials = rows$n_trials[1], img_size = rows$img_size[1], pairs = nrow(rows),
    mean_abs_d0 = round(mean(abs(rows$d0)), 4),
    mean_abs_dz = round(mean(abs(dz)), 4),
    q90_abs_dz = round(unname(quantile(abs(dz), 0.9)), 4),
    max_abs_dz = round(max(abs(dz)), 4),
    monte_carlo = round(mean(abs(monte_carlo(rows, cutoff))), 4)
  )
}

by_config <- split(results, list(results$n_trials, results$img_size), drop = TRUE)
knitr::kable(do.call(rbind, lapply(by_config, summary_row)), row.names = FALSE)
```

`monte_carlo` is the same quantity for a reference built from the same base image with a different
draw of the simulated responses. It is rerun-to-rerun variability, the difference between two
references of `iter` = `r formatC(iter, format = "d", big.mark = ",")` draws each, so it carries the
sampling error of both. What a published InfoVal carries is the error of *one* reference against the
population, which is smaller. That is measurable rather than assumable: scoring each 10,000-draw
reference against one of ten times the size, which stands in for the population.

```{r noise-yardsticks}
single_error <- function(rows, z) rows$single_shift + z * rows$single_rho

knitr::kable(signif(rbind(
  `the bug` = c(mean = mean(abs(shift(at_770, cutoff)))),
  `rerun to rerun` = c(mean = mean(abs(monte_carlo(at_770, cutoff)))),
  `one reference against the population` = c(mean = mean(abs(single_error(at_770, cutoff))))
), 3))
```

So at 770 trials and 64 pixels the bug moves an InfoVal about
`r sprintf("%.1f", mean(abs(shift(at_770, cutoff))) / mean(abs(single_error(at_770, cutoff))))`
times as far as the sampling error already in one, and about
`r sprintf("%.1f", mean(abs(shift(at_770, cutoff))) / mean(abs(monte_carlo(at_770, cutoff))))`
times a rerun of the same reference. The first is the comparison a reader interpreting a single
number wants; the second compares one difference with another.

The two terms contribute about equally at the cut-off: the median term averages
`r sprintf("%.4f", mean(abs(at_770$d0)))` and the scale term
`r sprintf("%.4f", cutoff * mean(abs(at_770$rho)))` once multiplied by 1.96. The scale term's closeness to its own Monte Carlo control
(`r sprintf("%.4f", mean(abs(at_770$rho)))` against `r sprintf("%.4f", mean(abs(at_770$mc_rho)))`)
invites reading it as mostly the error of estimating a MAD from `iter` draws, which a correctly
computed InfoVal carries too. The decomposition below measures that and finds otherwise; either way
the term is in the correction rather than absent from it.

```{r large-z}
at_z <- function(z, f) vapply(z, function(zz) mean(abs(f(at_770, zz))), numeric(1))

z <- c(2, 5, 10)
knitr::kable(data.frame(
  z = z,
  mean_abs_dz = round(at_z(z, shift), 4),
  as_share_of_z = round(at_z(z, shift) / z, 4),
  monte_carlo = round(at_z(z, monte_carlo), 4),
  ratio = round(at_z(z, shift) / at_z(z, monte_carlo), 2)
))
```

The margin over the Monte Carlo wobble narrows as the InfoVal grows, because both quantities pick up
the same scale term. At a large InfoVal the bug is barely distinguishable in size from re-running the
reference simulation with different random responses.

## Trials shrink it; resolution may enlarge it

```{r scaling}
by_trials <- do.call(rbind, lapply(split(at_64px, at_64px$n_trials), summary_row))
reference <- by_trials[by_trials$n_trials == 100, "mean_abs_d0"]
knitr::kable(data.frame(
  n_trials = by_trials$n_trials,
  observed = by_trials$mean_abs_d0,
  predicted_from_100 = round(reference * sqrt(100 / by_trials$n_trials), 4)
), row.names = FALSE)
```

The median term of the shift falls as one over the square root of the number of trials, which is what a reference
median built from an average over trials should do.

Pixels are a different story, and the answer is not the one a dimensional argument suggests. The
shift is a ratio of two quantities that both grow with resolution, so it might have been expected to
hold constant; measured at three sizes with the trial count fixed, its point estimate climbs.

```{r image-size}
by_size <- function(rows) {
  d0 <- abs(rows$d0); dz <- abs(shift(rows, cutoff))
  c(pairs = nrow(rows),
    mean_abs_d0 = mean(d0), se_d0 = sd(d0) / sqrt(nrow(rows)),
    mean_abs_dz = mean(dz), se_dz = sd(dz) / sqrt(nrow(rows)))
}

at_300 <- subset(results, n_trials == 300)
sizes <- do.call(rbind, lapply(split(at_300, at_300$img_size), by_size))
knitr::kable(signif(sizes, 3))
```

```{r image-size-test}
gap <- function(small, large, f) {
  x <- abs(f(subset(results, n_trials == 300 & img_size == small)))
  y <- abs(f(subset(results, n_trials == 300 & img_size == large)))
  se <- sqrt(var(x) / length(x) + var(y) / length(y))
  c(ratio = mean(y) / mean(x), difference = mean(y) - mean(x), se = se,
    ses_from_zero = abs(mean(y) - mean(x)) / se)
}

knitr::kable(signif(rbind(
  `d0, 512px vs 64px` = gap(64, 512, function(rows) rows$d0),
  `shift at the cut-off, 512px vs 64px` = gap(64, 512, function(rows) shift(rows, cutoff))
), 3))
```

At the default resolution the shift at the cut-off is
`r sprintf("%.2f", gap(64, 512, function(rows) shift(rows, cutoff))[["ratio"]])` times its value at
64px — but on `r nrow(subset(results, n_trials == 300 & img_size == 512))` pairs that is
`r sprintf("%.1f", gap(64, 512, function(rows) shift(rows, cutoff))[["ses_from_zero"]])` standard
errors, so it is neither established nor dismissed. Separating a ratio this size at three standard
errors means shrinking that ratio's standard error by a factor of three and a half, which at these
variances takes about twelve times the pairs in both groups — some 120 at 512 pixels and 240 at 64,
an order of magnitude beyond what this document runs.

Read the trial-count rows above as measured at 64px, then, and this 512px row as the one that applies
to a study at the package's own resolution: around
`r sprintf("%.2f", mean(abs(shift(subset(results, n_trials == 300 & img_size == 512), cutoff))))` in z
at 300 trials, against
`r sprintf("%.2f", mean(abs(shift(subset(results, n_trials == 300 & img_size == 64), cutoff))))` at
64px.

The configuration a real study actually runs — 512 pixels *and* several hundred trials — is measured
in neither series. Combining them requires the two effects to be separable, which is testable: the
trial-count ratio should be the same at either resolution.

```{r separability}
trial_ratio <- function(size) {
  at <- function(n) abs(shift(subset(results, n_trials == n & img_size == size), cutoff))
  x <- at(300); y <- at(770)
  c(pairs_300 = length(x), pairs_770 = length(y),
    ratio_770_over_300 = mean(y) / mean(x),
    se = (mean(y) / mean(x)) * sqrt(var(x) / (length(x) * mean(x)^2) +
                                    var(y) / (length(y) * mean(y)^2)))
}

knitr::kable(signif(rbind(`64px` = trial_ratio(64), `128px` = trial_ratio(128),
                          `sqrt(300/770)` = c(NA, NA, sqrt(300 / 770), NA)), 3))
```

The trial ratios come out above the square-root law at both resolutions, and the reason is visible in
the `monte_carlo` column of the earlier table, which does not fall with trials either. Only the
median term is an average over trials. Whether the scale term is mostly the error of estimating a
MAD from `iter` draws, or a real difference between the two bases' spreads, decides whether it can be
held fixed when combining the series — so it is separated rather than assumed, by differencing the
tenfold references against each other.

```{r scale-decomposition}
decomposed <- subset(results, !is.na(between_rho))
component <- function(x) c(mean = mean(abs(x)), se = sd(abs(x)) / sqrt(length(x)))

knitr::kable(signif(do.call(rbind, lapply(split(decomposed, decomposed$n_trials), function(rows)
  c(pairs = nrow(rows),
    median_term = component(rows$d0), between_base_median = component(rows$between_d0),
    scale_term = component(rows$rho), between_base_scale = component(rows$between_rho)))), 3))
```

Whether either between-base component falls with trials is what the combined estimate turns on, and
twenty and thirty pairs do not settle it from point estimates alone.

```{r decomposition-trend}
trend <- function(field) {
  x <- abs(decomposed[[field]][decomposed$n_trials == 300])
  y <- abs(decomposed[[field]][decomposed$n_trials == 770])
  difference <- mean(y) - mean(x)
  se <- sqrt(var(x) / length(x) + var(y) / length(y))
  c(at_300 = mean(x), at_770 = mean(y), difference = difference, se = se,
    ses_from_zero = abs(difference) / se)
}

knitr::kable(signif(rbind(`between-base median` = trend("between_d0"),
                          `between-base scale` = trend("between_rho")), 3))
```

A standard error against zero would not settle this: two references differ even when the thing they
estimate is identical, so `abs(between_rho)` has a positive expectation under the null of equal
spreads. The error has to be measured on this estimator rather than on a differently built stand-in,
so each pair estimates the same between-base quantity twice, from two shared draws. What the bases
really differ by is common to both estimates and cancels in their difference; what is left is the
estimator's own error.

```{r scale-against-null}
precise_770 <- decomposed[decomposed$n_trials == 770, ]

# Two estimates of one quantity: their average estimates it, their difference is
# error alone, and an error of that size on a single estimate is its sd over root two.
estimate <- (precise_770$between_rho + precise_770$between_rho_alt) / 2
error <- precise_770$between_rho - precise_770$between_rho_alt

c(between_base = mean(abs(estimate)), estimator_error = sd(error) / sqrt(2),
  ratio = mean(abs(estimate)) / (sd(error) / sqrt(2)),
  se_of_mean = sd(abs(estimate)) / sqrt(nrow(precise_770)))
```

Read that ratio before reading anything below it. Where the between-base magnitude stands well clear
of the estimator's own error, the two bases differ in spread and the scale term is not purely the
error of estimating a MAD — and reading `rho`'s resemblance to its Monte Carlo control as evidence
that it mostly was, as this document did earlier, went the wrong way. Where it does not, the earlier
reading survives and the term's composition is unsettled.

How much of the term is real is a further question this cannot answer. The 10,000-draw quantity is
the larger of the two, which is what estimation error on top of a real difference would produce, but
dividing one mean absolute value by the other does not give the share: taking absolute values
inflates whichever quantity carries more noise, and signed errors can cancel a real difference as
easily as add to it. That needs a variance model or repeated response seeds, and neither is here.

The two components behave differently across trial counts. The median one drops by
`r sprintf("%.1f", trend("between_d0")[["ses_from_zero"]])` standard errors, close to what the
square-root law predicts — `r sprintf("%.4f", trend("between_d0")[["at_300"]] * sqrt(300 / 770))`
against `r sprintf("%.4f", trend("between_d0")[["at_770"]])` observed. The scale one has a lower
point estimate at the higher count, but at
`r sprintf("%.1f", trend("between_rho")[["ses_from_zero"]])` standard errors these samples do not
separate that from no change, so whether it falls at all is open here.

Scaling the median term and holding the scale one fixed follows the trial law where it is measured
to hold and asserts nothing where it is not, which is the most these samples support. Which way that
errs is not determined: the scale term may shrink with trials and this treats none of it as
shrinking, but the two are signed, so carrying a larger one can cancel against the median as easily
as add to it.

```{r default-corner}
components <- function(rows) c(median_term = mean(abs(rows$d0)), scale_term = mean(abs(rows$rho)))
knitr::kable(signif(do.call(rbind, lapply(split(at_64px, at_64px$n_trials), components)), 3))

at_512 <- subset(results, img_size == 512)
trial_law <- sqrt(300 / 770)
estimate <- mean(abs(at_512$d0 * trial_law + cutoff * at_512$rho))
# Signed terms can cancel, so the estimate is not one-sided. Adding the two
# magnitudes instead cannot: per pair it bounds that pair, and averaged it bounds
# the mean rather than any individual study.
per_pair_bound <- abs(at_512$d0) * trial_law + cutoff * abs(at_512$rho)
bound <- mean(per_pair_bound)
c(measured_512px_300_trials = mean(abs(shift(at_512, cutoff))),
  estimated_512px_770_trials = estimate, mean_bound = bound,
  largest_pair_bound = max(per_pair_bound))
```

The median term falls by the square-root law; the scale term's own trend is not resolved above, so
the estimate scales the first and holds the second. That is a model rather than a bound: both terms are
signed and can cancel, so holding the scale term at its 300-trial value can as easily lower the
result as raise it. On that model a study at 512 pixels and 770 trials carries roughly
`r sprintf("%.2f", estimate)` in z, against
`r sprintf("%.2f", mean(abs(shift(at_512, cutoff))))` measured at 512 pixels and 300 trials — a
figure assembled from two measured components rather than measured in place. Adding the two
magnitudes rather than the signed terms removes the cancellation and gives
`r sprintf("%.2f", bound)`, which bounds the mean of these modelled shifts rather than any one of
them; the largest single pair among them reaches `r sprintf("%.2f", max(per_pair_bound))`.

This also puts the 24-trial reproduction in the issue in proportion. Extrapolating the square-root
law to 24 trials gives a median term of about `r sprintf("%.2f", reference * sqrt(100 / 24))`, and
the full shift there is not that number: no scale term is measured at 24 trials, and being signed it
can add to the median term or cancel against it. What the comparison supports is narrow: the −0.067
observed in the issue falls below the median term alone, so it is one draw rather than a measure of
the distribution it came from. The same law puts the median term for hundreds of trials well below
its 24-trial value — a statement about that component at 64 pixels, not about the full shift a real
study carries, which also holds an unmeasured scale term and whatever resolution does.

## What it takes to reach a different conclusion

A significance call at `z = 1.96` changes only when the correct InfoVal lies within the shift of the
cut-off *and* the shift points across it. Applying the measured shifts to a stand-in population of
true InfoVals:

```{r flip-rate}
# Each observation is reported as true_z + d0 + true_z * rho, so the scale term
# is evaluated at its own InfoVal rather than at the cut-off.
flip_rate <- function(rows, true_z) {
  reported <- outer(true_z, rows$d0, `+`) + outer(true_z, rows$rho)
  mean((true_z > cutoff) != (reported > cutoff))
}

set.seed(20260906)
spreads <- list("uniform 0-4" = runif(20000, 0, 4),
                "uniform 0-10" = runif(20000, 0, 10),
                "uniform 1.5-2.5" = runif(20000, 1.5, 2.5),
                "uniform 1.94-1.98" = runif(20000, 1.94, 1.98))
flipped <- vapply(spreads, function(z) flip_rate(at_770, z), numeric(1))
knitr::kable(data.frame(true_infoval_spread = names(spreads), flipped = round(flipped, 4)),
             row.names = FALSE)
```

Those spreads are stand-ins, not measurements: what InfoVals look like across real studies is not
something this package records, and none of them bounds the answer. The rate is set by how much of a
study's InfoVal distribution sits within the shift of the cut-off, so it falls as the distribution
broadens — `r sprintf("%.1f%%", 100 * flipped[["uniform 0-4"]])` over a four-unit spread,
`r sprintf("%.1f%%", 100 * flipped[["uniform 0-10"]])` over ten — and rises without a useful bound as
it narrows. The last row makes that concrete: concentrate every InfoVal within a few hundredths of
1.96 and `r sprintf("%.0f%%", 100 * flipped[["uniform 1.94-1.98"]])` of the calls change, approaching
the probability that the shift points across the threshold at all.

Which row a given study resembles is the whole question, and it is not one this package can answer:
nothing records what InfoVals look like in practice. A study whose classification images mostly carry
strong signal sits near the top rows; one that reports values hovering at the threshold sits near the
bottom one, and for it the rate is high. The effect of the bug is bounded only by where a study's
InfoVals actually fall.

And a classification image is only exposed if all of these hold:

1. InfoVal was computed at all — a feature from 0.4.0 onward, while CRAN carried 0.3.4.1 until the
   2021 archival.
2. The stimulus set has two or more base images.
3. It was generated with `use_same_parameters = FALSE`, which is not the default and appears in no
   vignette, example, test, or InfoVal configuration of the release gate.
4. The classification image comes from a base image other than the first — `(k-1)/k` of the images
   in such a study.

## What is not affected

`generateCI()` selects `stimuli_params[[baseimage]]`, so the classification images and z-maps
themselves are built from the right noise. The bug moves one number, the InfoVal, and only for base
images after the first.

## Limits

How often anyone sets `use_same_parameters = FALSE` is unknown. There is no telemetry, and the only
issue naming the setting is #299 itself, raised by an audit of the code rather than by anyone
reporting it from their own work — which is an absence of reports, not evidence of an absence of use.
Condition 3 above is therefore unquantified rather than measured to be rare. What this document
characterizes is the size of the damage to an affected classification image, in sampled means and
the largest value among the pairs it drew; it bounds neither that damage for an individual image nor
the number of affected studies.

The two references compared here share their simulated response draws, which reduces the Monte Carlo
error in their difference without removing it: 10,000 draws leave both medians and both MADs with
sampling error of their own, so `d0` and `rho` mix the noise realization with what remains of it.
That residual is not measured here. It is of the order of the single-reference error above, which
is smaller than the shift, and a fix drawing base 2's responses independently would add error of
that order again rather than leaving these numbers as they stand.

Resolution is measured at 300 trials, trial count at 64 and 128 pixels, and the corner a real study
occupies — 512 pixels with several hundred trials — in neither. The estimate for it is a product of
the two series. Its median term is scaled by a law measured to hold and its scale term is held flat,
which is a modelling choice in neither direction: the terms are signed, so holding the scale term up
can cancel the median term as easily as add to it. The resolution factor itself is not resolved at
all.

Everything reported under "the error is scatter" and "what it takes to reach a different conclusion"
is measured at 770 trials and 64 pixels. If the rise with resolution is real — and at under one
standard error this measurement neither shows it nor rules it out — those shifts and the flip rates
with them would be larger at 512 pixels; more trials would shrink the median term, while what they do
to the scale term is unresolved above, so the two do not simply cancel.

The shift measured here is the noise-realization component of what a fix corrects, measured with the
response draws held still across the two references. The package draws its responses from the state
left after stimulus regeneration, and a fix forwarding `use_same_parameters` consumes the later
bases' parameter draws first, so an old-versus-fixed comparison also moves the response sample, by
something of the order of the rerun column. With that said, this is the size to quote in a `NEWS.md`
"Reproducibility impact" entry: around 0.05 in z at 64 pixels and 770 trials, an estimated
0.06 for a default-sized study, and at the far end of the sampled pairs
`r sprintf("%.2f", max(abs(shift(results, cutoff))))` in the
`r results$n_trials[which.max(abs(shift(results, cutoff)))]`-trial configuration — all of them
shifts modelled for an InfoVal sitting at the cut-off, since this document generates no
classification images and so observes no movement of one. Scatter rather than a correction in a
direction; a median term falling as one over the square root of the trial count and a scale term
whose trend across trial counts these samples do not resolve; and no base image affected but the
second and later.
