---
title: "What the wrong-base InfoVal reference does to a conclusion"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "")
options(width = 100)
```

When stimuli are generated for several base images with `use_same_parameters = FALSE`, each base
gets its own parameter matrix, but `generateReferenceDistribution2IFC()` reconstructs the stimuli
without forwarding that setting. The null it builds is therefore always the first base image's
([issue #299](https://github.com/rdotsch/rcicr/issues/299)), and every base after the first is
scored against another base's noise.

This document measures what that costs. It exists because the two obvious readings are both wrong:
"the reference is from the wrong base, so the InfoVal is invalid" overstates it, and "it is only
noise, so it does not matter" understates it. The answer is that the wrong reference is a *valid*
null built on a different realization of the same noise process, so the error is scatter — with a
directional part some two orders of magnitude below it — whose size is set by the number of trials,
and a significance call flips only in a narrow window around the cut-off.

Knitting this runs the whole measurement, which takes tens of minutes: `r 30 + 20 + 20 + 12` base
image pairs, each needing two stimulus sets and three reference distributions of 10,000 iterations.

```{r constants}
library(rcicr)

img_size <- 64
nscales <- 5
iter <- 10000
response_seed <- 97
cutoff <- 1.96
```

```{r fixtures}
scratch <- tempfile("infoval-impact")
dir.create(scratch)

base_image <- function(name, seed, size) {
  f <- file.path(scratch, sprintf("%s-%d.png", name, size))
  set.seed(seed)
  png::writePNG(matrix(runif(size^2), size, size), f)
  f
}

# capture.output() here and below only keeps progress bars out of the rendered
# document; the package's functions are called exactly as a user would call them.
generate <- function(bases, n_trials, seed, size, same_parameters) {
  path <- tempfile("stim", tmpdir = scratch)
  dir.create(path)
  invisible(capture.output(
    generateStimuli2IFC(bases, n_trials = n_trials, img_size = size, stimulus_path = path,
                        seed = seed, use_same_parameters = same_parameters, nscales = nscales,
                        ncores = 1, save_as_png = FALSE)
  ))
  list.files(path, pattern = "\\.Rdata$", full.names = TRUE)[1]
}
```

The base images are synthetic noise rather than faces. Nothing here depends on their content: the
classification image and the reference norms are built from the noise alone, and the base image is
added only for display.

## The reference distribution ignores which base you score

Two base images, independent parameters, against the same stimulus set containing only the first of
them. If the reference carried any information about which base is being scored, these two runs
would differ.

```{r mechanism}
a <- base_image("a", 11, img_size)
b <- base_image("b", 12, img_size)

two_bases <- generate(list(a = a, b = b), n_trials = 24, seed = 31, size = img_size,
                      same_parameters = FALSE)
first_only <- generate(list(a = a), n_trials = 24, seed = 31, size = img_size,
                       same_parameters = TRUE)

# iter far below the recommended 10,000 here: this compares two runs against each
# other, not against a cut-off, and the warning is the function's own.
demo_reference <- function(rdata) {
  invisible(capture.output(norms <- suppressWarnings(generateReferenceDistribution2IFC(
    rdata, iter = 200, ncores = 1, response_seed = response_seed, save_rdata = FALSE))))
  norms
}

reference_two <- demo_reference(two_bases)
reference_one <- demo_reference(first_only)

saved <- new.env()
load(two_bases, envir = saved)

max(abs(saved$stimuli_params[["a"]] - saved$stimuli_params[["b"]]))
max(abs(reference_two - reference_one))
```

The two bases hold genuinely different parameters, and the reference distribution is bit-identical
to the one for a file that does not contain the second base at all.

## Base 2's own reference

The correction requires the null the second base should have had: its noise built from the
parameters actually saved for it. `generateNoiseImage()` does the reconstruction, and the loop below
is the arithmetic of `generateReferenceDistribution2IFC()`, batched so that the response draws are
consumed in the same order.

```{r glue}
noise_matrix <- function(params, p, size) {
  vapply(seq_len(nrow(params)),
         function(trial) as.vector(generateNoiseImage(params[trial, ], p)),
         numeric(size^2))
}

reference_norms <- function(noise, iter, seed, block = 500L) {
  set.seed(seed)
  n_trials <- ncol(noise)
  out <- numeric(iter)
  done <- 0L
  while (done < iter) {
    k <- min(block, iter - done)
    responses <- matrix(((runif(n_trials * k) > 0.5) * 2) - 1, nrow = n_trials, ncol = k)
    out[(done + 1):(done + k)] <- sqrt(colSums(((noise %*% responses) / n_trials)^2))
    done <- done + k
  }
  out
}
```

That reimplementation is only trustworthy if it reproduces the package's own output, so every
configuration below checks it against `generateReferenceDistribution2IFC()` on the first base before
using it on the second. The same check is what establishes that the *first* base is scored
correctly: the package's reference matches a null rebuilt from base 1's saved parameters.

```{r measure}
measure_pair <- function(n_trials, seed, size, validate) {
  rdata <- generate(list(a = base_image("a", 11, size), b = base_image("b", 12, size)),
                    n_trials = n_trials, seed = seed, size = size, same_parameters = FALSE)
  saved <- new.env()
  load(rdata, envir = saved)

  noise_a <- noise_matrix(saved$stimuli_params[["a"]], saved$p, size)
  noise_b <- noise_matrix(saved$stimuli_params[["b"]], saved$p, size)

  scored_against <- reference_norms(noise_a, iter, response_seed)   # what the bug uses
  own <- reference_norms(noise_b, iter, response_seed)              # what base 2 should have
  # Same base, an independent draw of the simulated responses: the Monte Carlo
  # wobble InfoVal already carries at this iter, as a yardstick for the shift.
  monte_carlo <- reference_norms(noise_a, iter, response_seed + 1L)

  agreement <- NA_real_
  if (validate) {
    invisible(capture.output(package_reference <- generateReferenceDistribution2IFC(
      rdata, iter = iter, ncores = 1, response_seed = response_seed, save_rdata = FALSE)))
    agreement <- max(abs(scored_against - package_reference))
    stopifnot(agreement < 1e-12)
  }

  data.frame(
    n_trials = n_trials, img_size = size, seed = seed, agreement = agreement,
    med_own = median(own), mad_own = mad(own),
    med_scored = median(scored_against), mad_scored = mad(scored_against),
    d0 = (median(own) - median(scored_against)) / mad(scored_against),
    rho = mad(own) / mad(scored_against) - 1,
    mc_shift = (median(monte_carlo) - median(scored_against)) / mad(scored_against),
    mc_rho = mad(monte_carlo) / mad(scored_against) - 1
  )
}

sweep <- function(n_trials, seeds, size = img_size) {
  do.call(rbind, Map(function(seed, validate) measure_pair(n_trials, seed, size, validate),
                     seeds, seq_along(seeds) == 1))
}
```

`d0` is the shift in the reference median, expressed in the units InfoVal is reported in, and `rho`
is the error in its scale. An InfoVal that should have been `z` is reported as `z + d0 + z * rho`,
so `d0` is what a threshold call sees and `rho` matters only for large values.

```{r sweep, results = "hide"}
results <- rbind(
  sweep(770, 1:30),
  sweep(300, 101:120),
  sweep(100, 201:220),
  sweep(300, 301:312, size = 128)
)
```

```{r agreement}
max(results$agreement, na.rm = TRUE)
```

```{r shift-definitions}
# An InfoVal that should read z is reported as z + shift(). The Monte Carlo
# yardstick has to carry the same scale term, or it is not the same quantity.
shift <- function(rows, z) rows$d0 + z * rows$rho
monte_carlo <- function(rows, z) rows$mc_shift + z * rows$mc_rho

centre <- function(x) {
  se <- sd(x) / sqrt(length(x))
  c(mean = mean(x), se = se, ses_from_zero = abs(mean(x)) / se)
}

at_64px <- subset(results, img_size == 64)
```

## The error is scatter, with a directional part too small to matter

```{r unbiased}
at_default <- subset(results, n_trials == 770)

knitr::kable(round(rbind(
  "reference median only (d0)" = centre(at_default$d0),
  "scale only (rho)" = centre(at_default$rho),
  "full shift at the cut-off" = centre(shift(at_default, cutoff)),
  "full shift at z = 10" = centre(shift(at_default, 10))
), 4))
```

Base 2's parameters are an independent draw from the same generator as base 1's, so the null it
receives is a legitimate one — just built on another realization of the noise. Across
`r nrow(at_default)` base image pairs at the package defaults the full shift at the cut-off averages
`r sprintf("%+.3f", mean(shift(at_default, cutoff)))`, which is
`r sprintf("%.1f", centre(shift(at_default, cutoff))[["ses_from_zero"]])` standard errors from zero
against a scatter `r sprintf("%.0f", sd(shift(at_default, cutoff)) / abs(mean(shift(at_default, cutoff))))`
times the size of that mean.

Neither term is exactly centred, though, and the sign of each is decided by the same structure.
`rho` is the ratio `mad(own) / mad(scored_against)` less one, and for exchangeable positive scales
`B/A + A/B >= 2` forces `E[B/A] >= 1`, strictly whenever the two differ at all. `d0` divides by the
MAD of one member of the pair only, so its mean is `-Cov(median, 1 / mad)` rather than zero: a noise
realization with more energy has both a larger median norm and a wider spread. Both terms therefore
carry a directional component, and the question is its size rather than its existence.

Neither mean can measure it — both sit far below their own standard errors. Averaging a pair's two
labellings can: had the second base image been listed first, the bug would have scored the first
against *its* null, and the two orderings share the bulk of the variation, which cancels.

```{r directional}
directional <- function(rows, z) {
  d0 <- ((rows$med_own - rows$med_scored) / rows$mad_scored +
         (rows$med_scored - rows$med_own) / rows$mad_own) / 2
  ratio <- 1 + rows$rho
  d0 + z * ((ratio + 1 / ratio) / 2 - 1)
}

knitr::kable(signif(do.call(rbind, lapply(split(at_64px, at_64px$n_trials), function(rows)
  c("median term" = centre(directional(rows, 0))[c("mean", "se")],
    "at the cut-off" = centre(directional(rows, cutoff))[c("mean", "se")],
    "at z = 10" = centre(directional(rows, 10))[c("mean", "se")]))), 3))
```

The estimates come out positive wherever they resolve against their own standard error, which is
what the argument above requires of the expectations, and all of them are minute. At the defaults the
directional part of the shift is
`r sprintf("%.1e", mean(directional(at_default, cutoff)))` at the cut-off and
`r sprintf("%.3f", mean(directional(at_default, 10)))` at an InfoVal of 10 — around
`r sprintf("%.0f%%", 100 * mean(directional(at_default, 10)) / mean(abs(shift(at_default, 10))))`
of the scatter at that value, a proportion that does not grow with the InfoVal because both scale
with it. A set of published InfoVals was therefore scattered rather than moved, to a precision far
finer than anyone reads these numbers to.

## How large the scatter is

```{r shift-table}
summary_row <- function(rows) {
  dz <- shift(rows, cutoff)
  data.frame(
    n_trials = rows$n_trials[1], img_size = rows$img_size[1], pairs = nrow(rows),
    mean_abs_d0 = round(mean(abs(rows$d0)), 4),
    mean_abs_dz = round(mean(abs(dz)), 4),
    q90_abs_dz = round(unname(quantile(abs(dz), 0.9)), 4),
    max_abs_dz = round(max(abs(dz)), 4),
    monte_carlo = round(mean(abs(monte_carlo(rows, cutoff))), 4)
  )
}

by_config <- split(results, list(results$n_trials, results$img_size), drop = TRUE)
knitr::kable(do.call(rbind, lapply(by_config, summary_row)), row.names = FALSE)
```

`monte_carlo` is the same quantity, computed the same way down to the scale term, for a reference
built from the same base image with a different draw of the simulated responses: the wobble any
InfoVal already carries at `iter` = `r formatC(iter, format = "d", big.mark = ",")`. At the default
`r max(results$n_trials)` trials the bug is worth
`r sprintf("%.1f", mean(abs(shift(at_default, cutoff))) / mean(abs(monte_carlo(at_default, cutoff))))`
times that wobble. It is the same kind of error as the one the method already tolerates, and not far
from the same size.

The scale error `rho` barely exceeds its own Monte Carlo control
(`r sprintf("%.4f", mean(abs(at_default$rho)))` against
`r sprintf("%.4f", mean(abs(at_default$mc_rho)))`), so almost all of the damage is the shift in the
median, and the `z * rho` term stays small until the InfoVal itself is large.

```{r large-z}
at_z <- function(z, f) vapply(z, function(zz) mean(abs(f(at_default, zz))), numeric(1))

z <- c(2, 5, 10)
knitr::kable(data.frame(
  z = z,
  mean_abs_dz = round(at_z(z, shift), 4),
  as_share_of_z = round(at_z(z, shift) / z, 4),
  monte_carlo = round(at_z(z, monte_carlo), 4),
  ratio = round(at_z(z, shift) / at_z(z, monte_carlo), 2)
))
```

The margin over the Monte Carlo wobble narrows as the InfoVal grows, because both quantities pick up
the same scale term. At a large InfoVal the bug is barely distinguishable in size from re-running the
reference simulation with different random responses.

## It is set by trials, not by pixels

```{r scaling}
by_trials <- do.call(rbind, lapply(split(at_64px, at_64px$n_trials), summary_row))
reference <- by_trials[by_trials$n_trials == 100, "mean_abs_d0"]
knitr::kable(data.frame(
  n_trials = by_trials$n_trials,
  observed = by_trials$mean_abs_d0,
  predicted_from_100 = round(reference * sqrt(100 / by_trials$n_trials), 4)
), row.names = FALSE)
```

The shift falls as one over the square root of the number of trials, which is what a reference
median built from an average over trials should do.

```{r image-size}
sizes <- lapply(c(64, 128), function(s) abs(subset(results, n_trials == 300 & img_size == s)$d0))
gap <- diff(vapply(sizes, mean, numeric(1)))
gap_se <- sqrt(sum(vapply(sizes, function(x) var(x) / length(x), numeric(1))))

c(`64px` = mean(sizes[[1]]), `128px` = mean(sizes[[2]]),
  difference = gap, se = gap_se, ses_from_zero = abs(gap) / gap_se)
```

Image size does not enter, then: at 300 trials the two differ by
`r sprintf("%.1f", abs(gap) / gap_se)` standard errors, so the 770-trial row above stands for a study
run at the default 512px, which is far too slow to measure directly. Sixteen times the pixels leaves
the shift where it was because both the reference median and its spread scale together, and the ratio
of the two is what InfoVal reports.

This also puts the 24-trial reproduction in the issue in proportion. Extrapolating the fitted
scaling to 24 trials gives a typical shift of
`r sprintf("%.2f", reference * sqrt(100 / 24))`, so the −0.067 observed there is one unremarkable
draw, and a real study with hundreds of trials sits several times below it.

## What it takes to reach a different conclusion

A significance call at `z = 1.96` changes only when the correct InfoVal lies within the shift of the
cut-off *and* the shift points across it. Applying the measured shifts to a stand-in population of
true InfoVals:

```{r flip-rate}
# Each observation is reported as true_z + d0 + true_z * rho, so the scale term
# is evaluated at its own InfoVal rather than at the cut-off.
flip_rate <- function(rows, true_z) {
  reported <- outer(true_z, rows$d0, `+`) + outer(true_z, rows$rho)
  mean((true_z > cutoff) != (reported > cutoff))
}

set.seed(20260906)
spreads <- list("uniform 0-4" = runif(20000, 0, 4),
                "uniform 0-10" = runif(20000, 0, 10),
                "uniform 1.5-2.5" = runif(20000, 1.5, 2.5),
                "uniform 1.94-1.98" = runif(20000, 1.94, 1.98))
flipped <- vapply(spreads, function(z) flip_rate(at_default, z), numeric(1))
knitr::kable(data.frame(true_infoval_spread = names(spreads), flipped = round(flipped, 4)),
             row.names = FALSE)
```

Those spreads are stand-ins, not measurements: what InfoVals look like across real studies is not
something this package records, and none of them bounds the answer. The rate is set by how much of a
study's InfoVal distribution sits within the shift of the cut-off, so it falls as the distribution
broadens — `r sprintf("%.1f%%", 100 * flipped[["uniform 0-4"]])` over a four-unit spread,
`r sprintf("%.1f%%", 100 * flipped[["uniform 0-10"]])` over ten — and rises without a useful bound as
it narrows. The last row makes that concrete: concentrate every InfoVal within a few hundredths of
1.96 and `r sprintf("%.0f%%", 100 * flipped[["uniform 1.94-1.98"]])` of the calls change, approaching
the probability that the shift points across the threshold at all. The reassurance is that real
InfoVals are not distributed that way, not that the bug's effect is bounded.

And a classification image is only exposed if all of these hold:

1. InfoVal was computed at all — a feature from 0.4.0 onward, while CRAN carried 0.3.4.1 until the
   2021 archival.
2. The stimulus set has two or more base images.
3. It was generated with `use_same_parameters = FALSE`, which is not the default and appears in no
   vignette, example, test, or InfoVal configuration of the release gate.
4. The classification image comes from a base image other than the first — `(k-1)/k` of the images
   in such a study.

## What is not affected

`generateCI()` selects `stimuli_params[[baseimage]]`, so the classification images and z-maps
themselves are built from the right noise. The bug moves one number, the InfoVal, and only for base
images after the first.

## Limits

How often anyone sets `use_same_parameters = FALSE` is unknown. There is no telemetry, and no issue
in the tracker has ever mentioned the setting, so condition 3 above is unquantified rather than
measured to be rare — this document bounds the damage per affected classification image, not the
number of affected studies.

The two references compared here share their simulated response draws, which isolates the noise
realization from Monte Carlo error. A fix that draws base 2's responses from its own position in the
stream adds roughly the `monte_carlo` column on top, which changes none of the conclusions above.

The shift measured here is the correction a fix applies, so it is also the size to quote in a
`NEWS.md` "Reproducibility impact" entry: about
`r sprintf("%.2f", mean(abs(shift(at_default, cutoff))))` in z at the defaults, scaling as one over
the square root of the trial count, scatter rather than a correction in a direction, and affecting
no base image but the second and later.
