---
title: "What the wrong-base InfoVal reference does to a conclusion"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "")
options(width = 100)
```

When stimuli are generated for several base images with `use_same_parameters = FALSE`, each base
gets its own parameter matrix, but `generateReferenceDistribution2IFC()` reconstructs the stimuli
without forwarding that setting. The null it builds is therefore always the first base image's
([issue #299](https://github.com/rdotsch/rcicr/issues/299)), and every base after the first is
scored against another base's noise.

This document characterizes the discrepancy using synthetic stimulus sets. A reference from the
same generator is not necessarily calibrated for another base's saved noise. The measured shifts
vary across pairs, with a small estimated directional component in the configurations examined.
The median component decreases across the sampled trial counts; the scale component's trend is
unresolved. Neither the prevalence of affected studies nor their changes in significance calls is
measured here.

Knitting runs the full sweep of `r 30 + 20 + 20 + 12 + 8 + 10` base image pairs, including ten at
512 pixels. Every pair uses three 10,000-draw references; the fifty 64px pairs at 300 or 770 trials
also use four 100,000-draw references. Runtime depends on the available hardware.

```{r constants}
library(rcicr)

img_size <- 64
nscales <- 5
iter <- 10000
response_seed <- 97
cutoff <- 1.96
```

```{r fixtures}
scratch <- tempfile("infoval-impact")
dir.create(scratch)

base_image <- function(name, seed, size) {
  f <- file.path(scratch, sprintf("%s-%d.png", name, size))
  set.seed(seed)
  png::writePNG(matrix(runif(size^2), size, size), f)
  f
}

# capture.output() here and below only keeps progress bars out of the rendered
# document; the package's functions are called exactly as a user would call them.
# ?generateReferenceDistribution2IFC guarantees the stimuli, and so everything
# measured here, do not depend on ncores. mclapply() below rejects more than one
# core on Windows, which knits there serially rather than not at all.
cores <- if (.Platform$OS.type == "windows") 1L else max(1L, parallel::detectCores() - 1L)

generate <- function(bases, n_trials, seed, size, same_parameters) {
  path <- tempfile("stim", tmpdir = scratch)
  dir.create(path)
  invisible(capture.output(
    generateStimuli2IFC(bases, n_trials = n_trials, img_size = size, stimulus_path = path,
                        seed = seed, use_same_parameters = same_parameters, nscales = nscales,
                        ncores = cores, save_as_png = FALSE)
  ))
  list.files(path, pattern = "\\.Rdata$", full.names = TRUE)[1]
}
```

The base images are synthetic noise rather than faces. Nothing here depends on their content: the
classification image and the reference norms are built from the noise alone, and the base image is
added only for display.

## The reference distribution ignores which base you score

Two base images, independent parameters, against the same stimulus set containing only the first of
them. If the reference carried any information about which base is being scored, these two runs
would differ.

```{r mechanism}
a <- base_image("a", 11, img_size)
b <- base_image("b", 12, img_size)

two_bases <- generate(list(a = a, b = b), n_trials = 24, seed = 31, size = img_size,
                      same_parameters = FALSE)
first_only <- generate(list(a = a), n_trials = 24, seed = 31, size = img_size,
                       same_parameters = TRUE)

# iter far below the recommended 10,000 here: this compares two runs against each
# other, not against a cut-off, and the warning is the function's own.
demo_reference <- function(rdata) {
  invisible(capture.output(norms <- suppressWarnings(generateReferenceDistribution2IFC(
    rdata, iter = 200, ncores = 1, response_seed = response_seed, save_rdata = FALSE))))
  norms
}

reference_two <- demo_reference(two_bases)
reference_one <- demo_reference(first_only)

saved <- new.env()
load(two_bases, envir = saved)

max(abs(saved$stimuli_params[["a"]] - saved$stimuli_params[["b"]]))
max(abs(reference_two - reference_one))
```

The two bases hold genuinely different parameters, and the reference distribution is bit-identical
to the one for a file that does not contain the second base at all.

## Base 2's own reference

The correction requires the null the second base should have had: its noise built from the
parameters actually saved for it. `generateNoiseImage()` does the reconstruction, and the loop below
is the arithmetic of `generateReferenceDistribution2IFC()`, batched so that the response draws are
consumed in the same order.

```{r glue}
noise_matrix <- function(params, p, size) {
  matrix(unlist(parallel::mclapply(seq_len(nrow(params)), mc.cores = cores,
                function(trial) as.vector(generateNoiseImage(params[trial, ], p)))),
         nrow = size^2)
}

# A block holds one column per iteration, so its width comes from the image size,
# keeping the intermediate near 200 MB at any resolution.
reference_norms <- function(noise, iter, seed, block = max(50L, as.integer(2.5e7 / nrow(noise)))) {
  set.seed(seed)
  n_trials <- ncol(noise)
  out <- numeric(iter)
  done <- 0L
  while (done < iter) {
    k <- min(block, iter - done)
    responses <- matrix(((runif(n_trials * k) > 0.5) * 2) - 1, nrow = n_trials, ncol = k)
    out[(done + 1):(done + k)] <- sqrt(colSums(((noise %*% responses) / n_trials)^2))
    done <- done + k
  }
  out
}
```

The first pair of each configuration below 512px checks this reimplementation against
`generateReferenceDistribution2IFC()` on the first base. At 512px that comparison is skipped because
of the repeated matrix conversion in #306. Agreement in the checked configurations supports the
reimplementation and the first-base reconstruction; it is not a direct validation at 512px.

```{r measure}
measure_pair <- function(n_trials, seed, size, validate) {
  rdata <- generate(list(a = base_image("a", 11, size), b = base_image("b", 12, size)),
                    n_trials = n_trials, seed = seed, size = size, same_parameters = FALSE)
  saved <- new.env()
  load(rdata, envir = saved)

  noise_a <- noise_matrix(saved$stimuli_params[["a"]], saved$p, size)
  noise_b <- noise_matrix(saved$stimuli_params[["b"]], saved$p, size)

  scored_against <- reference_norms(noise_a, iter, response_seed)   # what the bug uses
  own <- reference_norms(noise_b, iter, response_seed)              # what base 2 should have
  # Same base, an independent draw of the simulated responses: the Monte Carlo
  # wobble InfoVal already carries at this iter, as a yardstick for the shift.
  monte_carlo <- reference_norms(noise_a, iter, response_seed + 1L)
  # Finite-reference proxy for the population, restricted to the configurations
  # used by the comparison and trend diagnostics below.
  decompose <- size == 64 && n_trials %in% c(300, 770)
  precise <- if (decompose) reference_norms(noise_a, 10L * iter, response_seed + 2L) else NULL
  # Match response draws across bases to retain the main comparison's coupling.
  # Finite-reference error remains at this larger iteration count.
  precise_b <- if (decompose) reference_norms(noise_b, 10L * iter, response_seed + 2L) else NULL
  # Replicate the same between-base estimator with a second shared response draw.
  # Differencing removes its common target and bias, not just its population target.
  precise_alt <- if (decompose) reference_norms(noise_a, 10L * iter, response_seed + 5L) else NULL
  precise_b_alt <- if (decompose) reference_norms(noise_b, 10L * iter, response_seed + 5L) else NULL

  agreement <- NA_real_
  if (validate) {
    invisible(capture.output(package_reference <- generateReferenceDistribution2IFC(
      rdata, iter = iter, ncores = 1, response_seed = response_seed, save_rdata = FALSE)))
    agreement <- max(abs(scored_against - package_reference))
    stopifnot(agreement < 1e-12)
  }

  data.frame(
    n_trials = n_trials, img_size = size, seed = seed, agreement = agreement,
    med_own = median(own), mad_own = mad(own),
    med_scored = median(scored_against), mad_scored = mad(scored_against),
    d0 = (median(own) - median(scored_against)) / mad(scored_against),
    rho = mad(own) / mad(scored_against) - 1,
    mc_shift = (median(monte_carlo) - median(scored_against)) / mad(scored_against),
    mc_rho = mad(monte_carlo) / mad(scored_against) - 1,
    # Oriented like d0 and rho above: the reference that defines the InfoVal on
    # top, the one actually used underneath, so the two are comparable at a z
    # meaning the same thing.
    single_shift = if (is.null(precise)) NA_real_ else
      (median(precise) - median(scored_against)) / mad(scored_against),
    single_rho = if (is.null(precise)) NA_real_ else mad(precise) / mad(scored_against) - 1,
    between_d0 = if (is.null(precise_b)) NA_real_ else
      (median(precise_b) - median(precise)) / mad(precise),
    between_rho = if (is.null(precise_b)) NA_real_ else mad(precise_b) / mad(precise) - 1,
    between_d0_alt = if (is.null(precise_alt)) NA_real_ else
      (median(precise_b_alt) - median(precise_alt)) / mad(precise_alt),
    between_rho_alt = if (is.null(precise_alt)) NA_real_ else
      mad(precise_b_alt) / mad(precise_alt) - 1
  )
}

# Validated on the first pair of each configuration, except at 512px, where
# generateReferenceDistribution2IFC() rebuilds a matrix from its stimulus frame on
# every one of the 10,000 iterations (#306). That frame carries 64 times the pixels
# it does at 64px, where the check is already the slowest thing in this document,
# so the cost there is prohibitive. The reimplementation does not vary with
# resolution.
sweep <- function(n_trials, seeds, size = img_size) {
  do.call(rbind, Map(function(seed, validate) measure_pair(n_trials, seed, size, validate),
                     seeds, seq_along(seeds) == 1 & size < 512))
}
```

`d0` is the shift in the reference median, expressed in the units InfoVal is reported in, and `rho`
is the error in its scale. An InfoVal that should have been `z` is reported as `z + d0 + z * rho`,
so both terms reach a threshold call — `rho` multiplied by the threshold — and the scale term grows
in weight as the InfoVal does.

```{r sweep, results = "hide"}
results <- rbind(
  sweep(770, 1:30),
  sweep(300, 101:120),
  sweep(100, 201:220),
  sweep(300, 301:312, size = 128),
  sweep(770, 501:508, size = 128),
  sweep(300, 401:410, size = 512)
)
```

```{r agreement}
max(results$agreement, na.rm = TRUE)
```

```{r shift-definitions}
# An InfoVal that should read z is reported as z + shift(). The Monte Carlo
# yardstick has to carry the same scale term, or it is not the same quantity.
shift <- function(rows, z) rows$d0 + z * rows$rho
monte_carlo <- function(rows, z) rows$mc_shift + z * rows$mc_rho

centre <- function(x) {
  se <- sd(x) / sqrt(length(x))
  c(mean = mean(x), se = se, ses_from_zero = abs(mean(x)) / se)
}

at_64px <- subset(results, img_size == 64)
```

Standard errors across base pairs describe variation conditional on the selected response streams.
The same response seeds are reused across pairs, and response seeds are not independently resampled
for each row. These standard errors do not include all uncertainty over response simulations.

## Scatter and the estimated directional component

```{r unbiased}
# 770 trials at 64 pixels. Not the package's default configuration, whose
# 512 pixels are measured separately below and at fewer trials.
at_770 <- subset(results, n_trials == 770 & img_size == 64)

knitr::kable(round(rbind(
  "reference median only (d0)" = centre(at_770$d0),
  "scale only (rho)" = centre(at_770$rho),
  "full shift at the cut-off" = centre(shift(at_770, cutoff)),
  "full shift at z = 10" = centre(shift(at_770, 10))
), 4))
```

Base 2's parameters come from the same generator as base 1's, but its reference should be
conditional on its own saved noise. Across
`r nrow(at_770)` base image pairs at 770 trials and 64 pixels the full shift at the cut-off averages
`r sprintf("%+.3f", mean(shift(at_770, cutoff)))`, which is
`r sprintf("%.1f", centre(shift(at_770, cutoff))[["ses_from_zero"]])` standard errors from zero
against a scatter `r sprintf("%.0f", sd(shift(at_770, cutoff)) / abs(mean(shift(at_770, cutoff))))`
times the size of that mean.

The scale ratio need not be centred at zero.
`rho` is the ratio `mad(own) / mad(scored_against)` less one, and for exchangeable positive scales
`B/A + A/B >= 2` forces `E[B/A] >= 1`, strictly when they differ with positive probability
(assuming finite expectations). This inequality does not determine the sign of the median term:
`d0` also depends on how the two medians relate to the denominator MAD. Shared response draws can
couple the two reference estimates, so cross-base dependence cannot simply be dropped.

Averaging each pair's two labellings gives a symmetric estimator of the directional component
under exchangeability. It uses both possible orderings of the same pair, with the standard error
computed across pairs rather than treating the two orderings as independent observations.

```{r directional}
directional <- function(rows, z) {
  d0 <- ((rows$med_own - rows$med_scored) / rows$mad_scored +
         (rows$med_scored - rows$med_own) / rows$mad_own) / 2
  ratio <- 1 + rows$rho
  d0 + z * ((ratio + 1 / ratio) / 2 - 1)
}

knitr::kable(signif(do.call(rbind, lapply(split(at_64px, at_64px$n_trials), function(rows)
  c("median term" = centre(directional(rows, 0))[c("mean", "se")],
    "at the cut-off" = centre(directional(rows, cutoff))[c("mean", "se")],
    "at z = 10" = centre(directional(rows, 10))[c("mean", "se")]))), 3))
```

At 770 trials and 64 pixels the estimated directional part of the shift is
`r sprintf("%.1e", mean(directional(at_770, cutoff)))` at the cut-off and
`r sprintf("%.3f", mean(directional(at_770, 10)))` at an InfoVal of 10 — around
`r sprintf("%.0f%%", 100 * mean(directional(at_770, 10)) / mean(abs(shift(at_770, 10))))`
of the mean absolute shift at that value. These are estimates for the sampled configurations and
the specified InfoVals, not evidence that directional effects are absent or irrelevant in every
study. The tables show their uncertainty alongside their magnitude.

## How large the scatter is

```{r shift-table}
summary_row <- function(rows) {
  dz <- shift(rows, cutoff)
  data.frame(
    n_trials = rows$n_trials[1], img_size = rows$img_size[1], pairs = nrow(rows),
    mean_abs_d0 = round(mean(abs(rows$d0)), 4),
    mean_abs_dz = round(mean(abs(dz)), 4),
    q90_abs_dz = round(unname(quantile(abs(dz), 0.9)), 4),
    max_abs_dz = round(max(abs(dz)), 4),
    monte_carlo = round(mean(abs(monte_carlo(rows, cutoff))), 4)
  )
}

by_config <- split(results, list(results$n_trials, results$img_size), drop = TRUE)
knitr::kable(do.call(rbind, lapply(by_config, summary_row)), row.names = FALSE)
```

`monte_carlo` is the same quantity for a reference built from the same base image with a different
draw of the simulated responses. It is rerun-to-rerun variability, the difference between two
references of `iter` = `r formatC(iter, format = "d", big.mark = ",")` draws each, so it carries the
sampling error of both. Error against an exact population reference is a different quantity.
The next comparison approximates it with a 100,000-draw reference, which still has estimation error;
the table's population label denotes this finite-reference proxy.

```{r noise-yardsticks}
single_error <- function(rows, z) rows$single_shift + z * rows$single_rho

knitr::kable(signif(rbind(
  `the bug` = c(mean = mean(abs(shift(at_770, cutoff)))),
  `rerun to rerun` = c(mean = mean(abs(monte_carlo(at_770, cutoff)))),
  `one reference against the population` = c(mean = mean(abs(single_error(at_770, cutoff))))
), 3))
```

So at 770 trials and 64 pixels the bug moves an InfoVal about
`r sprintf("%.1f", mean(abs(shift(at_770, cutoff))) / mean(abs(single_error(at_770, cutoff))))`
times as far as the discrepancy against that finite-reference proxy, and about
`r sprintf("%.1f", mean(abs(shift(at_770, cutoff))) / mean(abs(monte_carlo(at_770, cutoff))))`
times the rerun-to-rerun discrepancy. Neither comparison measures the exact population error.

The two terms contribute about equally at the cut-off: the median term averages
`r sprintf("%.4f", mean(abs(at_770$d0)))` and the scale term
`r sprintf("%.4f", cutoff * mean(abs(at_770$rho)))` once multiplied by 1.96. The scale term's closeness to its own Monte Carlo control
(`r sprintf("%.4f", mean(abs(at_770$rho)))` against `r sprintf("%.4f", mean(abs(at_770$mc_rho)))`)
does not establish how much comes from estimation error and how much from a difference in the
underlying reference spreads. The diagnostics below explore this without assigning component shares.

```{r large-z}
at_z <- function(z, f) vapply(z, function(zz) mean(abs(f(at_770, zz))), numeric(1))

z <- c(2, 5, 10)
knitr::kable(data.frame(
  z = z,
  mean_abs_dz = round(at_z(z, shift), 4),
  as_share_of_z = round(at_z(z, shift) / z, 4),
  monte_carlo = round(at_z(z, monte_carlo), 4),
  ratio = round(at_z(z, shift) / at_z(z, monte_carlo), 2)
))
```

In these sampled rows the ratio narrows at the displayed larger InfoVals. Both discrepancies
include a scale component multiplied by z, but those components are different random quantities.
Similarity of their mean magnitudes is not an equivalence test.

## Trial count and resolution

```{r scaling}
by_trials <- do.call(rbind, lapply(split(at_64px, at_64px$n_trials), summary_row))
reference <- by_trials[by_trials$n_trials == 100, "mean_abs_d0"]
knitr::kable(data.frame(
  n_trials = by_trials$n_trials,
  observed = by_trials$mean_abs_d0,
  predicted_from_100 = round(reference * sqrt(100 / by_trials$n_trials), 4)
), row.names = FALSE)
```

The sampled mean absolute median component is close to the displayed inverse-square-root curve.
This is a descriptive comparison at 64px, not an established law for every trial count or resolution.

Pixels are a different story, and the answer is not the one a dimensional argument suggests. The
shift is a ratio of two quantities that both grow with resolution, so it might have been expected to
hold constant; measured at three sizes with the trial count fixed, its point estimate climbs.

```{r image-size}
by_size <- function(rows) {
  d0 <- abs(rows$d0); dz <- abs(shift(rows, cutoff))
  c(pairs = nrow(rows),
    mean_abs_d0 = mean(d0), se_d0 = sd(d0) / sqrt(nrow(rows)),
    mean_abs_dz = mean(dz), se_dz = sd(dz) / sqrt(nrow(rows)))
}

at_300 <- subset(results, n_trials == 300)
sizes <- do.call(rbind, lapply(split(at_300, at_300$img_size), by_size))
knitr::kable(signif(sizes, 3))
```

```{r image-size-test}
gap <- function(small, large, f) {
  x <- abs(f(subset(results, n_trials == 300 & img_size == small)))
  y <- abs(f(subset(results, n_trials == 300 & img_size == large)))
  se <- sqrt(var(x) / length(x) + var(y) / length(y))
  c(ratio = mean(y) / mean(x), difference = mean(y) - mean(x), se = se,
    ses_from_zero = abs(mean(y) - mean(x)) / se)
}

knitr::kable(signif(rbind(
  `d0, 512px vs 64px` = gap(64, 512, function(rows) rows$d0),
  `shift at the cut-off, 512px vs 64px` = gap(64, 512, function(rows) shift(rows, cutoff))
), 3))
```

At the default resolution the shift at the cut-off is
`r sprintf("%.2f", gap(64, 512, function(rows) shift(rows, cutoff))[["ratio"]])` times its value at
64px — but on `r nrow(subset(results, n_trials == 300 & img_size == 512))` pairs that is
`r sprintf("%.1f", gap(64, 512, function(rows) shift(rows, cutoff))[["ses_from_zero"]])` standard
errors for the difference of means, so a resolution effect remains unresolved. The ratio is a point
estimate, not a measured resolution multiplier that can be applied to other configurations.

Read the trial-count rows above as measured at 64px, then, and this 512px row as the one that applies
to a study at the package's own resolution: around
`r sprintf("%.2f", mean(abs(shift(subset(results, n_trials == 300 & img_size == 512), cutoff))))` in z
at 300 trials, against
`r sprintf("%.2f", mean(abs(shift(subset(results, n_trials == 300 & img_size == 64), cutoff))))` at
64px.

The specific combination of 512 pixels and 770 trials is not measured. A separable model would
predict the same trial-count ratio at each resolution; the next table describes those ratios at
64px and 128px but does not establish separability at 512px.

```{r separability}
trial_ratio <- function(size) {
  at <- function(n) abs(shift(subset(results, n_trials == n & img_size == size), cutoff))
  x <- at(300); y <- at(770)
  c(pairs_300 = length(x), pairs_770 = length(y),
    ratio_770_over_300 = mean(y) / mean(x),
    se = (mean(y) / mean(x)) * sqrt(var(x) / (length(x) * mean(x)^2) +
                                    var(y) / (length(y) * mean(y)^2)))
}

knitr::kable(signif(rbind(`64px` = trial_ratio(64), `128px` = trial_ratio(128),
                          `sqrt(300/770)` = c(NA, NA, sqrt(300 / 770), NA)), 3))
```

The point estimates for the full shift differ from the inverse-square-root prediction. The shift
contains both median and scale components, each with finite-reference estimation error. Comparing
the tenfold references gives further diagnostics, but neither establishes a scale trend nor
justifies holding it fixed when extrapolating.

```{r scale-decomposition}
decomposed <- subset(results, !is.na(between_rho))
component <- function(x) c(mean = mean(abs(x)), se = sd(abs(x)) / sqrt(length(x)))

knitr::kable(signif(do.call(rbind, lapply(split(decomposed, decomposed$n_trials), function(rows)
  c(pairs = nrow(rows),
    median_term = component(rows$d0), between_base_median = component(rows$between_d0),
    scale_term = component(rows$rho), between_base_scale = component(rows$between_rho)))), 3))
```

Whether either between-base component falls with trials is what the combined estimate turns on, and
twenty and thirty pairs do not settle it from point estimates alone.

```{r decomposition-trend}
trend <- function(field) {
  x <- abs(decomposed[[field]][decomposed$n_trials == 300])
  y <- abs(decomposed[[field]][decomposed$n_trials == 770])
  difference <- mean(y) - mean(x)
  se <- sqrt(var(x) / length(x) + var(y) / length(y))
  c(at_300 = mean(x), at_770 = mean(y), difference = difference, se = se,
    ses_from_zero = abs(difference) / se)
}

knitr::kable(signif(rbind(`between-base median` = trend("between_d0"),
                          `between-base scale` = trend("between_rho")), 3))
```

A positive mean absolute difference alone cannot establish unequal population spreads: finite
references differ even under equality. Each pair therefore repeats the between-base estimate with
a second response seed. Conditional on the saved bases, differencing the replicates removes their
common target and any common bias. It probes response-sampling variability, not bias relative to
the population MAD ratio.

```{r scale-against-null}
precise_770 <- decomposed[decomposed$n_trials == 770, ]

# Under conditionally iid replicates, sd(error) / sqrt(2) describes single-replicate
# noise; the average has noise scale sd(error) / 2. Neither measures common bias.
estimate <- (precise_770$between_rho + precise_770$between_rho_alt) / 2
error <- precise_770$between_rho - precise_770$between_rho_alt

c(between_base = mean(abs(estimate)), estimator_error = sd(error) / sqrt(2),
  ratio = mean(abs(estimate)) / (sd(error) / sqrt(2)),
  se_of_mean = sd(abs(estimate)) / sqrt(nrow(precise_770)))
```

This ratio is descriptive, not a calibrated test of equal population spreads. Its numerator is
the mean absolute value of a two-replicate average; its denominator estimates the standard deviation
of a single replicate's sampling error, assuming independent, identically distributed response
replicates conditional on each base pair. For the average, the corresponding noise scale would be
`sd(error) / 2`. Mean absolute values and standard deviations are different summaries, and neither
comparison removes common finite-sample ratio bias. Sharing response seeds across base pairs also
means that the across-pair summaries do not capture all response-seed uncertainty.

The component shares and the existence of a population spread difference remain unresolved by this
diagnostic. Establishing them requires a specified inferential model and uncertainty calculation;
an arbitrary threshold on the displayed ratio cannot substitute for that. No conclusion below
depends on treating this ratio as a hypothesis test.

In these samples the median component's mean absolute value drops by
`r sprintf("%.1f", trend("between_d0")[["ses_from_zero"]])` standard errors, close to what the
inverse-square-root model predicts — `r sprintf("%.4f", trend("between_d0")[["at_300"]] * sqrt(300 / 770))`
against `r sprintf("%.4f", trend("between_d0")[["at_770"]])` observed. The scale one has a lower
point estimate at the higher count, but at
`r sprintf("%.1f", trend("between_rho")[["ses_from_zero"]])` standard errors these samples do not
separate that from no change, so whether it falls at all is open here.

The following extrapolation assumes inverse-square-root scaling for the median component and no
change in the scale component of each sampled 512px pair. Both are modelling assumptions at the
unmeasured trial count. An unresolved scale trend does not support a zero trend, and the direction
of error is unknown because signed components can cancel.

```{r default-corner}
components <- function(rows) c(median_term = mean(abs(rows$d0)), scale_term = mean(abs(rows$rho)))
knitr::kable(signif(do.call(rbind, lapply(split(at_64px, at_64px$n_trials), components)), 3))

at_512 <- subset(results, img_size == 512)
trial_law <- sqrt(300 / 770)
estimate <- mean(abs(at_512$d0 * trial_law + cutoff * at_512$rho))
# Signed terms can cancel, so the estimate is not one-sided. Adding the two
# magnitudes instead cannot: per pair it bounds that pair, and averaged it bounds
# the mean rather than any individual study.
per_pair_bound <- abs(at_512$d0) * trial_law + cutoff * abs(at_512$rho)
bound <- mean(per_pair_bound)
c(measured_512px_300_trials = mean(abs(shift(at_512, cutoff))),
  estimated_512px_770_trials = estimate, mean_bound = bound,
  largest_pair_bound = max(per_pair_bound))
```

Under these assumptions the estimate scales the median term and holds the scale term fixed.
That is a model rather than a bound on real studies: both terms are
signed and can cancel, so holding the scale term at its 300-trial value can as easily lower the
result as raise it. On that model a study at 512 pixels and 770 trials carries roughly
`r sprintf("%.2f", estimate)` in z, against
`r sprintf("%.2f", mean(abs(shift(at_512, cutoff))))` measured at 512 pixels and 300 trials — a
modelled sample mean rather than a measurement at that configuration. Adding the two
magnitudes rather than the signed terms removes the cancellation and gives
`r sprintf("%.2f", bound)`, which bounds the mean of these modelled shifts rather than any one of
them; the largest single pair among them reaches `r sprintf("%.2f", max(per_pair_bound))`.

Extrapolating the 64px inverse-square-root model to 24 trials gives a mean absolute median
component of about `r sprintf("%.2f", reference * sqrt(100 / 24))`. That is not an estimate of the
full shift in #299: the issue used 512 reference draws and a particular synthetic CI, whereas the
sweep uses 10,000 draws and models shifts at the cut-off. No 24-trial scale component is measured
here. The issue's single reproduction demonstrates the bug, not its typical impact.

## What it takes to reach a different conclusion

A significance call at `z = 1.96` changes only when the correct InfoVal lies within the shift of the
cut-off *and* the shift points across it. Applying the measured shifts to a stand-in population of
true InfoVals:

```{r flip-rate}
# Each observation is reported as true_z + d0 + true_z * rho, so the scale term
# is evaluated at its own InfoVal rather than at the cut-off.
flip_rate <- function(rows, true_z) {
  reported <- outer(true_z, rows$d0, `+`) + outer(true_z, rows$rho)
  mean((true_z > cutoff) != (reported > cutoff))
}

set.seed(20260906)
spreads <- list("uniform 0-4" = runif(20000, 0, 4),
                "uniform 0-10" = runif(20000, 0, 10),
                "uniform 1.5-2.5" = runif(20000, 1.5, 2.5),
                "uniform 1.94-1.98" = runif(20000, 1.94, 1.98))
flipped <- vapply(spreads, function(z) flip_rate(at_770, z), numeric(1))
knitr::kable(data.frame(true_infoval_spread = names(spreads), flipped = round(flipped, 4)),
             row.names = FALSE)
```

These are artificial populations crossed independently with every sampled reference pair. In
these examples the flip rate is `r sprintf("%.1f%%", 100 * flipped[["uniform 0-4"]])` for the
four-unit uniform spread and `r sprintf("%.1f%%", 100 * flipped[["uniform 0-10"]])` for the
ten-unit spread, rising to `r sprintf("%.0f%%", 100 * flipped[["uniform 1.94-1.98"]])` in the
near-threshold example. This is not a general monotonic relationship with distribution width:
location, shift direction, and dependence between InfoVals and reference pairs also matter.
None of these examples estimates an affected-study rate or bounds one beyond the trivial 0–100%.

And a classification image is only exposed if all of these hold:

1. InfoVal was computed at all — a feature from 0.4.0 onward, while CRAN carried 0.3.4.1 until the
   2021 archival.
2. The stimulus set has two or more base images.
3. It was generated with the non-default `use_same_parameters = FALSE`.
4. The classification image comes from a base image other than the first. For k bases, k-1
   bases are exposed; the fraction of analyzed CIs depends on how many are computed for each base.

## What is not affected

`generateCI()` selects `stimuli_params[[baseimage]]`, so the classification images and z-maps
themselves are built from the right noise. The bug moves one number, the InfoVal, and only for base
images after the first.

## Limits

How often anyone sets `use_same_parameters = FALSE` is unknown. Issue #299 was raised by a code
audit and does not provide usage data from researchers' studies.
Condition 3 above is therefore unquantified rather than measured to be rare. What this document
characterizes is the size of the damage to an affected classification image, in sampled means and
the largest value among the pairs it drew; it bounds neither that damage for an individual image nor
the number of affected studies.

The references share simulated response draws. Their medians and MADs retain sampling error,
so the measured shift combines differences between saved noise realizations with finite-reference
error under that coupling. Sharing draws can change the variance of the difference; its direction
and size cannot be inferred from the single-reference control. The repeated tenfold references
diagnose variability at their own iteration count, not the residual error in the 10,000-draw shift.

Resolution comparisons use 300 trials; trial-count comparisons use 64px and 128px. The 512px,
770-trial result is an extrapolation with unresolved resolution dependence and scale trend.
The threshold illustrations use 770 trials at 64px and cannot be transferred to 512px merely by
multiplying by a mean shift ratio. Flip rates depend on the joint distribution of InfoVals and
shifts, not only on mean absolute shift. The artificial populations assume independence from the
sampled reference pairs; actual study responses need not satisfy that assumption.

This is not an old-versus-fixed package comparison. A fix must select the requested base and
separate cached references; forwarding `use_same_parameters` alone is insufficient (#299).
Its RNG behavior depends on the implementation and whether `response_seed` is supplied. The
eventual fix needs its own reproducibility measurement before quoting a correction in `NEWS.md`.

For orientation, the shared-response mean absolute shift at the cut-off is
`r sprintf("%.2f", mean(abs(shift(at_770, cutoff))))` at 64px and 770 trials, with
`r sprintf("%.2f", estimate)` under the 512px, 770-trial extrapolation. The largest sampled value is
`r sprintf("%.2f", max(abs(shift(results, cutoff))))` in the
`r results$n_trials[which.max(abs(shift(results, cutoff)))]`-trial configuration — all of them
shifts modelled for an InfoVal sitting at the cut-off, since this document generates no
classification images and so observes no movement of one. Sample means and maxima bound neither
individual studies nor the population distribution of shifts.
