---
title: "What the individual-CI mislabelling did to a second-stage analysis"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "")
options(width = 100)
```

The bug fixed in rcicr 1.3.0 permuted output filenames: the file named `unique(participants)[i]` held the classification image of `sort(unique(participants))[i]`. The [advisory](https://rdotsch.github.io/rcicr/articles/rcicr-individual-ci-advisory.html) tells researchers what to do about it. This document is the evidence behind the claims it makes, and exists because those claims are easy to get wrong in either direction: "it only adds noise" and "it could have produced spurious findings" are each half right in ways that matter.

```{r constants}
n_simulations <- 20000
alpha <- 0.05
set.seed(20260819)
```

## The permutation the bug applied

For each position in the participants vector, which participant's image did the file at that position actually hold? `generateCI()` looped over the sorted levels and named the output from the order of appearance, so the two orders have to be lined up against each other.

```{r permutation}
misfiled_as <- function(participant_ids) {
  appearance_order <- unique(participant_ids)
  loop_order <- sort(unique(participant_ids))
  match(loop_order, appearance_order)
}
```

`sort()` is correct for every input type: it follows a factor's own levels, sorts numbers numerically, and sorts character strings by the session's collation, the same three behaviours `factor()` gave the original run.

## How much of the pairing survives

With identifiers `p1 … pN` in collection order (the ordinary labelling scheme), almost nothing does, once there are ten or more participants.

```{r survival}
survival <- data.frame(
  identifiers = c(sprintf("p1..p%d", c(9, 12, 20, 30, 50, 99, 100, 120)),
                  "p01..p12 (zero-padded)"),
  stringsAsFactors = FALSE
)
id_sets <- c(lapply(c(9, 12, 20, 30, 50, 99, 100, 120),
                    function(n) paste0("p", seq_len(n))),
             list(sprintf("p%02d", 1:12)))
survival$participants <- lengths(id_sets)
survival$kept_own_image <- vapply(id_sets, function(ids) {
  sum(misfiled_as(ids) == seq_along(unique(ids)))
}, numeric(1))
survival$share_kept <- round(survival$kept_own_image / survival$participants, 3)
knitr::kable(survival)
```

This is not monotone in N because the count turns on where the identifiers change width. At 99, 100 and 101 participants eleven files keep their own image (at N = 100 they are `p1` and `p80`–`p89`), while at 120 it is back to one.

## The designs

Four designs are simulated, plus a sweep over how scrambled the ordering was. Each pairs a participant-level variable with the images and asks what the mislabelling did to the analysis.

```{r correlational-design}
simulate_correlational_design <- function(n_participants, true_correlation,
                                          draw_covariate,
                                          iterations = n_simulations) {
  misfiled <- misfiled_as(paste0("p", seq_len(n_participants)))
  detected_correct <- detected_mislabelled <- logical(iterations)
  significant_opposite <- logical(iterations)
  correlation_recovered <- numeric(iterations)

  for (i in seq_len(iterations)) {
    covariate <- draw_covariate(n_participants)
    outcome <- true_correlation * covariate +
      sqrt(1 - true_correlation^2) * rnorm(n_participants)

    test_correct <- cor.test(covariate, outcome)
    test_mislabelled <- cor.test(covariate, outcome[misfiled])

    detected_correct[i] <- test_correct$p.value < alpha
    detected_mislabelled[i] <- test_mislabelled$p.value < alpha
    significant_opposite[i] <- test_mislabelled$p.value < alpha &&
      test_mislabelled$estimate < 0
    correlation_recovered[i] <- test_mislabelled$estimate
  }

  # Both analyses see the same outcome vector each iteration, so the paired
  # difference is the statistic that carries the claim: its error is far
  # smaller than either rate's, and it does not confound the permutation with
  # the test's own size at small N. Its SE comes from the per-iteration signed
  # differences rather than from sqrt(discordant)/iterations, which is the
  # McNemar SE *under equal rates* and overstates the error wherever the two
  # rates genuinely differ.
  decisions_changed <- sum(detected_correct != detected_mislabelled)
  signed_difference <- detected_mislabelled - detected_correct
  data.frame(
    N = n_participants,
    true_correlation = true_correlation,
    detected_correct = mean(detected_correct),
    detected_mislabelled = mean(detected_mislabelled),
    difference = mean(detected_mislabelled) - mean(detected_correct),
    difference_se = sd(signed_difference) / sqrt(iterations),
    decisions_changed = decisions_changed / iterations,
    became_significant = sum(!detected_correct & detected_mislabelled) / iterations,
    mean_correlation_recovered = mean(correlation_recovered),
    significant_opposite = mean(significant_opposite)
  )
}
```

### A: a covariate unrelated to labelling order

The ordinary individual-differences design: raters judge each classification image, and those judgments are related back to something about the participant (their condition, their score on another measure, their group) that has nothing to do with the order participants happened to be numbered in.

```{r scenario-a}
grid <- expand.grid(N = c(12, 20, 30, 50), true_correlation = c(0, 0.3, 0.5))
scenario_a <- do.call(rbind, Map(
  function(n, rho) simulate_correlational_design(n, rho, function(k) rnorm(k)),
  grid$N, grid$true_correlation
))
knitr::kable(scenario_a, digits = 3)
```

### B: a covariate that *is* collection order

Testing date, cohort and session number all advance as data are collected, which is exactly what the identifiers do too.

```{r scenario-b}
scenario_b <- do.call(rbind, Map(
  function(n, rho) simulate_correlational_design(
    n, rho, function(k) as.numeric(scale(seq_len(k)))
  ),
  grid$N, grid$true_correlation
))
knitr::kable(scenario_b, digits = 3)
```

### C: condition assigned in blocks by collection order

```{r scenario-c}
simulate_blocked_design <- function(n_participants, effect_size,
                                    iterations = n_simulations) {
  misfiled <- misfiled_as(paste0("p", seq_len(n_participants)))
  condition <- rep(c(0, 1), each = n_participants / 2)
  detected_correct <- detected_mislabelled <- logical(iterations)
  significant_predicted <- significant_opposite <- logical(iterations)

  for (i in seq_len(iterations)) {
    outcome <- effect_size * condition + rnorm(n_participants)
    test_correct <- t.test(outcome[condition == 1], outcome[condition == 0])

    mislabelled_outcome <- outcome[misfiled]
    test_mislabelled <- t.test(mislabelled_outcome[condition == 1],
                               mislabelled_outcome[condition == 0])
    observed_difference <- test_mislabelled$estimate[1] - test_mislabelled$estimate[2]

    detected_correct[i] <- test_correct$p.value < alpha
    detected_mislabelled[i] <- test_mislabelled$p.value < alpha
    significant_predicted[i] <- test_mislabelled$p.value < alpha &&
      observed_difference > 0
    significant_opposite[i] <- test_mislabelled$p.value < alpha &&
      observed_difference < 0
  }

  decisions_changed <- sum(detected_correct != detected_mislabelled)
  signed_difference <- detected_mislabelled - detected_correct
  data.frame(
    N = n_participants,
    effect_size = effect_size,
    detected_correct = mean(detected_correct),
    detected_mislabelled = mean(detected_mislabelled),
    difference = mean(detected_mislabelled) - mean(detected_correct),
    difference_se = sd(signed_difference) / sqrt(iterations),
    decisions_changed = decisions_changed / iterations,
    became_significant = sum(!detected_correct & detected_mislabelled) / iterations,
    significant_predicted = mean(significant_predicted),
    significant_opposite = mean(significant_opposite),
    crossed_condition = mean(condition[misfiled] != condition)
  )
}

blocked_grid <- expand.grid(N = c(12, 20, 30, 50), effect_size = c(0, 0.8))
scenario_c <- do.call(rbind, Map(
  function(n, d) simulate_blocked_design(n, d),
  blocked_grid$N, blocked_grid$effect_size
))
knitr::kable(scenario_c, digits = 3)
```

## Finding 1: the false positive rate is unchanged

Where there was no association to begin with, rejection stays at the nominal 5% in every design and at every sample size.

```{r null-cells}
shared_columns <- c("N", "detected_correct", "detected_mislabelled",
                    "difference", "difference_se", "decisions_changed",
                    "became_significant")
null_cells <- rbind(
  data.frame(design = "A", test = "cor.test",
             scenario_a[scenario_a$true_correlation == 0, shared_columns]),
  data.frame(design = "B", test = "cor.test",
             scenario_b[scenario_b$true_correlation == 0, shared_columns]),
  data.frame(design = "C", test = "Welch t",
             scenario_c[scenario_c$effect_size == 0, shared_columns])
)
null_cells$SEs_from_zero <- ifelse(
  null_cells$difference_se > 0,
  abs(null_cells$difference) / null_cells$difference_se, NA
)
knitr::kable(null_cells, digits = 3, row.names = FALSE)
```

Do not read the individual rates too closely: the Monte Carlo SE at 0.05 is `r sprintf("%.4f", sqrt(0.05 * 0.95 / n_simulations))` at this many iterations, and each test brings its own size: the Welch *t* in design C is mildly conservative at N = 12, in the correctly labelled column too. The statistic that carries the claim is the **paired** difference, both analyses seeing the same outcome vector within an iteration. It runs from `r sprintf("%+.3f", min(null_cells$difference))` to `r sprintf("%+.3f", max(null_cells$difference))`, each within `r sprintf("%.1f", max(null_cells$SEs_from_zero, na.rm = TRUE))` paired SEs of zero.

That result is structural rather than a lucky set of draws, *for an outcome that is exchangeable across participants*: permuting one side of a pair that is independent of the other leaves it independent, so the null distribution is unchanged by construction. The null cells here are iid Gaussian and satisfy that. Independence alone is not enough: if the outcome's distribution shifts along collection order, say because its variance grows as data collection went on, then the permutation moves high-leverage observations against the covariate and a Pearson or Welch test's rejection rate can shift even with no true association. Nothing in these runs speaks to that case.

**A dataset can still get a different verdict even when the aggregate rate is unchanged.** The two analyses reach a different verdict on `r sprintf("%.1f%%", 100 * min(null_cells$decisions_changed))` to `r sprintf("%.1f%%", 100 * max(null_cells$decisions_changed))` of null datasets, of which `r sprintf("%.0f%%", 100 * min(null_cells$became_significant / null_cells$decisions_changed, na.rm = TRUE))` to `r sprintf("%.0f%%", 100 * max(null_cells$became_significant / null_cells$decisions_changed, na.rm = TRUE))` are rejections the correct labels would not have given. A result that became significant only on mislabelled files is a false positive the bug handed over, however untouched the aggregate rate is.

## Finding 2: a real association is weakened in proportion to the scrambling

In design A the association is lost, and under the `p1 … pN` scheme that loss is close to total: at N = 50 a true correlation of 0.5 goes from `r sprintf("%.3f", scenario_a$detected_correct[scenario_a$N == 50 & scenario_a$true_correlation == 0.5])` detected to `r sprintf("%.3f", scenario_a$detected_mislabelled[scenario_a$N == 50 & scenario_a$true_correlation == 0.5])`. The runs that then fail to reject are the false negatives.

How much is lost depends on how far the ordering departed from sorted, which scenario E sweeps using the real mechanism: build an appearance order, let `misfiled_as()` derive the permutation from it.

```{r scenario-e}
simulate_severity <- function(n_participants, true_correlation, pairs_swapped,
                              iterations = n_simulations) {
  identifiers <- sprintf("p%02d", seq_len(n_participants))
  appearance_order <- identifiers
  if (pairs_swapped > 0) {
    first_of_pair <- seq(1, by = 2, length.out = pairs_swapped)
    appearance_order[c(rbind(first_of_pair, first_of_pair + 1))] <-
      appearance_order[c(rbind(first_of_pair + 1, first_of_pair))]
  }
  misfiled <- misfiled_as(appearance_order)

  detected_correct <- detected_mislabelled <- logical(iterations)
  for (i in seq_len(iterations)) {
    covariate <- rnorm(n_participants)
    outcome <- true_correlation * covariate +
      sqrt(1 - true_correlation^2) * rnorm(n_participants)
    detected_correct[i] <- cor.test(covariate, outcome)$p.value < alpha
    detected_mislabelled[i] <-
      cor.test(covariate, outcome[misfiled])$p.value < alpha
  }

  data.frame(
    pairs_swapped = pairs_swapped,
    share_correctly_paired = mean(misfiled == seq_len(n_participants)),
    detected_correct = mean(detected_correct),
    detected_mislabelled = mean(detected_mislabelled)
  )
}

scenario_e <- do.call(rbind, lapply(
  c(1, 2, 5, 10, 25),
  function(swaps) simulate_severity(50, 0.5, swaps)
))
knitr::kable(scenario_e, digits = 3)
```

Detection tracks the share still correctly paired almost one-for-one until the pairing is nearly gone. Two participants entered out of sequence cost a couple of points of power; the `p1 … pN` scheme at 50 participants keeps `r sprintf("%.2f", 1 / 50)` of the pairing and is the bottom of that range.

The share fixes only the average attenuation. The permutation's *cycle structure* separately moves the variance, so two orderings keeping the same share can differ in detection. Comparing disjoint transpositions against a single long cycle displacing the same number of participants:

```{r cycle-structure}
detection_for <- function(misfiled, n_participants = 50, true_correlation = 0.5,
                          iterations = n_simulations) {
  detected <- logical(iterations)
  for (i in seq_len(iterations)) {
    covariate <- rnorm(n_participants)
    outcome <- true_correlation * covariate +
      sqrt(1 - true_correlation^2) * rnorm(n_participants)
    detected[i] <- cor.test(covariate, outcome[misfiled])$p.value < alpha
  }
  mean(detected)
}

as_transpositions <- function(n_participants, displaced) {
  misfiled <- seq_len(n_participants)
  first_of_pair <- seq(1, by = 2, length.out = displaced / 2)
  misfiled[c(rbind(first_of_pair, first_of_pair + 1))] <-
    misfiled[c(rbind(first_of_pair + 1, first_of_pair))]
  misfiled
}

as_one_long_cycle <- function(n_participants, displaced) {
  misfiled <- seq_len(n_participants)
  cycle <- seq_len(displaced)
  misfiled[cycle] <- c(cycle[-1], cycle[1])
  misfiled
}

cycle_structure <- do.call(rbind, lapply(c(10, 20, 40, 50), function(displaced) {
  transposed <- as_transpositions(50, displaced)
  cycled <- as_one_long_cycle(50, displaced)
  data.frame(
    displaced = displaced,
    share_correctly_paired = mean(transposed == seq_len(50)),
    detected_transpositions = detection_for(transposed),
    detected_one_long_cycle = detection_for(cycled)
  )
}))
knitr::kable(cycle_structure, digits = 3, row.names = FALSE)
```

The gap reaches `r sprintf("%.3f", max(abs(cycle_structure$detected_one_long_cycle - cycle_structure$detected_transpositions)))` at the severe end. So these rows are worked examples rather than a lookup table.

## Finding 3: where something tracks collection order, the residual takes either sign

In designs B and C the permutation is no longer arbitrary with respect to the design. A `d = 0.8` condition difference survives attenuated and correctly signed at N = 50, where a quarter of participants cross the condition boundary, and comes back **reversed** at N = 20, where four fifths do.

## Finding 4: the shuffle can also raise apparent support

Design D asks whether the permutation can hand back a hypothesis-consistent result rather than only destroy one. The assignment scheme decides the answer, so both are run: contiguous blocks already maximise the trend contrast under correct labels, so any permutation can only reduce it, and testing that alone would make "the shuffle never helps" true by construction.

```{r scenario-d}
simulate_trend_design <- function(n_participants, trend, assignment,
                                  iterations = n_simulations) {
  misfiled <- misfiled_as(paste0("p", seq_len(n_participants)))
  condition <- if (assignment == "blocked") {
    rep(c(0, 1), each = n_participants / 2)
  } else {
    rep(c(0, 1), times = n_participants / 2)
  }
  collection_order <- as.numeric(scale(seq_len(n_participants)))

  correct_predicted <- mislabelled_predicted <- logical(iterations)
  mislabelled_opposite <- logical(iterations)

  for (i in seq_len(iterations)) {
    outcome <- trend * collection_order + rnorm(n_participants)
    test_correct <- t.test(outcome[condition == 1], outcome[condition == 0])
    mislabelled_outcome <- outcome[misfiled]
    test_mislabelled <- t.test(mislabelled_outcome[condition == 1],
                               mislabelled_outcome[condition == 0])

    effect_correct <- test_correct$estimate[1] - test_correct$estimate[2]
    effect_mislabelled <- test_mislabelled$estimate[1] - test_mislabelled$estimate[2]

    correct_predicted[i] <- test_correct$p.value < alpha && effect_correct > 0
    mislabelled_predicted[i] <- test_mislabelled$p.value < alpha &&
      effect_mislabelled > 0
    mislabelled_opposite[i] <- test_mislabelled$p.value < alpha &&
      effect_mislabelled < 0
  }

  data.frame(
    N = n_participants,
    trend = trend,
    assignment = assignment,
    correct_predicted = mean(correct_predicted),
    mislabelled_predicted = mean(mislabelled_predicted),
    mislabelled_opposite = mean(mislabelled_opposite),
    amplified = mean(mislabelled_predicted) > mean(correct_predicted)
  )
}

trend_grid <- expand.grid(N = c(12, 20, 30, 50), trend = c(0.3, 0.6, 1.0),
                          assignment = c("blocked", "alternating"),
                          stringsAsFactors = FALSE)
scenario_d <- do.call(rbind, Map(
  function(n, t, a) simulate_trend_design(n, t, a),
  trend_grid$N, trend_grid$trend, trend_grid$assignment
))
knitr::kable(scenario_d, digits = 3, row.names = FALSE)
```

Alternating assignment is the opposite case: the correct pairing spreads the trend evenly across conditions, leaving little contrast to destroy and room to create one. The mislabelled analysis raises hypothesis-consistent rejections above the correctly labelled one in `r sum(scenario_d$amplified[scenario_d$assignment == "alternating"])` of `r sum(scenario_d$assignment == "alternating")` alternating cells.

**So the mislabelling can hand back apparent support.** What it cannot do is know which direction anyone predicted: the sign it pushes is fixed by the identifier scheme and the assignment scheme together, and nothing in sorting refers to a hypothesis. How often that sign nevertheless agrees with a given prediction is *not* answered here: every cell fixes the trend and the "predicted" direction as positive rather than sampling them, so these runs cannot support a claim about directional inflation across a literature.

## Checking one specific analysis

The advisory's one-line check settles whether an ordering was affected. This goes further and estimates what it cost, for a given participant vector and an assumed effect size. It takes the vector itself, so factors, numbers and character strings all behave as they did in the original run.

```{r checker}
ordering_impact <- function(participants, true_correlation = 0.5,
                            iterations = 2000) {
  # Validated before anything is allocated: iterations = 0 would return NaN for
  # every advertised rate, and a non-numeric one would fail inside logical()
  # with an implementation-level error rather than a usable message.
  if (length(iterations) != 1 || !is.finite(iterations) || iterations < 1 ||
        iterations != trunc(iterations)) {
    stop("iterations must be a whole number of at least 1", call. = FALSE)
  }
  if (length(true_correlation) != 1 || !is.finite(true_correlation) ||
        abs(true_correlation) > 1) {
    stop("true_correlation must be a correlation between -1 and 1", call. = FALSE)
  }
  misfiled <- misfiled_as(participants)
  n_participants <- length(unique(participants))
  share_correctly_paired <- mean(misfiled == seq_len(n_participants))

  # The permutation is derived with sort(), which for character identifiers
  # uses this session's collation and not necessarily the one the original
  # analysis ran under. That makes every result here -- including an
  # all-clear -- conditional, so the condition travels in the return value
  # rather than in prose somebody can drop. Comparing against C order, which
  # method = "radix" gives regardless of LC_COLLATE, is evidence where the two
  # disagree; agreement is not proof, since a third locale may collate letter
  # pairs as units (Czech "ch" after "h", Danish "aa" last).
  # Every type carries a condition, because for every type sort() answers a
  # question about the object it was given rather than about the original run.
  # A factor is the sharpest case: its levels *are* the loop order, so
  # factor(c("a","b"), levels = c("b","a")) is fully swapped while a rebuilt
  # factor(c("a","b")) reports a clean all-clear.
  distinct_ids <- unique(participants)
  conditions <- if (is.factor(participants)) {
    c(sprintf("ordering follows this factor's levels: %s",
              paste(levels(participants), collapse = ", ")),
      "assumes those are the levels the original participants vector carried")
  } else if (is.character(participants)) {
    differs_from_c_order <- !identical(sort(distinct_ids),
                                       sort(distinct_ids, method = "radix"))
    c(sprintf("ordering derived under LC_COLLATE=%s%s",
              Sys.getlocale("LC_COLLATE"),
              if (differs_from_c_order) {
                "; differs from C order, so the original locale decides"
              } else {
                "; identical under C order"
              }),
      "assumes participants is the original object, not a character copy")
  } else {
    "assumes participants was numeric originally, not the same digits as text"
  }

  if (share_correctly_paired == 1) {
    return(list(share_correctly_paired = 1, affected = FALSE,
                conditional_on = conditions))
  }
  if (n_participants < 3) {
    return(list(share_correctly_paired = share_correctly_paired,
                affected = TRUE, conditional_on = conditions,
                note = "too few participants to estimate rates"))
  }

  detected_correct <- detected_mislabelled <- logical(iterations)
  for (i in seq_len(iterations)) {
    covariate <- rnorm(n_participants)
    outcome <- true_correlation * covariate +
      sqrt(1 - true_correlation^2) * rnorm(n_participants)
    detected_correct[i] <- cor.test(covariate, outcome)$p.value < alpha
    detected_mislabelled[i] <-
      cor.test(covariate, outcome[misfiled])$p.value < alpha
  }
  list(share_correctly_paired = share_correctly_paired,
       affected = TRUE,
       conditional_on = conditions,
       detected_correct = mean(detected_correct),
       detected_mislabelled = mean(detected_mislabelled),
       decisions_changed = mean(detected_correct != detected_mislabelled))
}

str(ordering_impact(paste0("p", 1:12), iterations = 2000))

try(ordering_impact(paste0("p", 1:12), iterations = 0))
```

A permutation that only exchanges participants sharing a value on the second-stage predictor leaves that analysis exactly as computed, however wrong the filenames are. The advisory gives readers `identical(predictor, predictor[misfiled_as(participants)])` for that, and it is worth ruling in first because it can settle the question outright, though the ordinary `p1 … pN` scheme with conditions in blocks does not survive it.

The implication runs one way only. `TRUE` proves the analysis is untouched; `FALSE` does not prove it changed, since two swapped participants whose images scored alike would also leave it as it was. What settles a `FALSE` is repairing the filenames and re-running, which the advisory asks for anyway.

The detection rates assume a Pearson correlation on normally distributed data, so treat them as indicative if the second stage was a condition comparison or a regression. `share_correctly_paired` is exact, but exact for the permutation *this session* derives, which for character identifiers depends on its collation. That is why the result carries `conditional_on` rather than leaving the caveat in prose: an all-clear from the wrong locale is still an all-clear to whoever reads the return value.

```{r checker-conditions}
str(ordering_impact(c("d", "ch"), iterations = 500))

# a factor's levels are the loop order, so a rebuilt one answers a different
# question from the original -- the condition says which levels were used
str(ordering_impact(factor(c("a", "b"), levels = c("b", "a")), iterations = 500))
```

Lowercase ASCII is no guarantee here, since some locales collate letter pairs as units. Czech, for instance, orders `ch` after `h`, and traditional Danish puts `aa` at the end of the alphabet, so `c("d", "ch")` is sorted under Czech collation and swapped under C. Run the check under the original locale wherever the identifiers are anything more than digits and simple prefixes.
