---
title: "What a parallel worker costs"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "")
options(width = 100)
```

Every parallel loop in rcicr starts worker processes, and each holds its own copy of the noise basis plus the working memory of one render. This document measures what that costs at 512 pixels, on `main` and on the change for [#405](https://github.com/rdotsch/rcicr/issues/405):

1. renders use the basis with its patch indices as integer, while the `.Rdata` file keeps the basis as built;
2. in parallel, each participant's rows travel with that participant's task;
3. no more workers start than the loop has tasks.

Each configuration runs once, in a fresh R process started in a session of its own. Memory is the peak resident set size summed over that session's processes, workers included, sampled every 0.1 s; it counts pages shared between processes once per process. The machine and R version are printed below; absolute times and sizes are this machine's.

## Two installed versions

```{r versions}
repo <- normalizePath("..")
main_commit <- system2("git", c("-C", shQuote(repo), "rev-parse", "origin/main"), stdout = TRUE)
branch_commit <- system2("git", c("-C", shQuote(repo), "rev-parse", "HEAD"), stdout = TRUE)
install <- function(commit) {
  source_dir <- tempfile("src")
  lib <- tempfile("lib")
  dir.create(source_dir)
  dir.create(lib)
  system2("sh", c("-c", shQuote(sprintf("git -C %s archive %s | tar -x -C %s", shQuote(repo), commit,
                                        shQuote(source_dir)))))
  status <- system2(file.path(R.home("bin"), "R"),
                    c("CMD", "INSTALL", "--no-test-load", "-l", shQuote(lib), shQuote(source_dir)),
                    stdout = FALSE, stderr = FALSE)
  stopifnot(status == 0)
  lib
}
libs <- c(main = install(main_commit), branch = install(branch_commit))
c(main = main_commit, branch = branch_commit)
c(R = R.version.string, cores = parallel::detectCores(),
  memory_GB = round(as.numeric(sub("\\D+(\\d+).*", "\\1",
                                   grep("MemTotal", readLines("/proc/meminfo"), value = TRUE))) / 2^20))
```

## The measurement

`job.R` runs one call and saves its result, so every result can be compared with `identical()`. `measure.sh` samples memory while it runs.

```{r scripts}
job <- tempfile(fileext = ".R")
writeLines(c(
  'suppressMessages(library(rcicr))',
  'a <- commandArgs(TRUE); what <- a[1]; rdata <- a[2]; cores <- as.integer(a[3]); result <- a[4]',
  'n <- as.integer(a[5])',
  'quiet <- function(e) { utils::capture.output(v <- suppressMessages(e)); v }',
  'set.seed(9)',
  'out <- tempfile(); dir.create(out)',
  'elapsed <- system.time(value <- switch(what,',
  '  participants = {',
  '    per <- 1400 %/% n',
  '    quiet(generateCI(rep(seq_len(per), n), sample(c(-1, 1), per * n, TRUE), "face", rdata,',
  '                     participants = rep(seq_len(n), each = per), save_as_png = FALSE,',
  '                     n_cores = cores))',
  '  },',
  '  stimuli = {',
  '    quiet(generateStimuli2IFC(list(face = rdata), n_trials = n, img_size = 512, stimulus_path = out,',
  '                              seed = 1, ncores = cores, save_as_png = FALSE, save_rdata = FALSE,',
  '                              return_as_dataframe = TRUE))',
  '  },',
  '  zmap = {',
  '    quiet(generateCI(seq_len(n), sample(c(-1, 1), n, TRUE), "face", rdata, save_as_png = FALSE,',
  '                     zmap = TRUE, zmapmethod = "t.test", zmapdecoration = FALSE,',
  '                     zmaptargetpath = out, n_cores = cores))$zmap',
  '  },',
  '  reference = {',
  '    quiet(generateReferenceDistribution2IFC(rdata, iter = 100, ncores = cores, response_seed = 1,',
  '                                            save_rdata = FALSE, reference_method = "images"))',
  '  }))[["elapsed"]]',
  'saveRDS(value, result)',
  'cat(elapsed, "\\n")'
), job)

sampler <- tempfile(fileext = ".sh")
writeLines(c(
  '#!/bin/bash',
  'out=$(mktemp)',
  'setsid Rscript "$@" > "$out" 2>/dev/null &',
  'pid=$!',
  'peak=0',
  'while kill -0 $pid 2>/dev/null; do',
  '  s=$(ps -o rss= --sid $pid 2>/dev/null | awk \'{t+=$1} END {print int(t/1024)}\')',
  '  [ "${s:-0}" -gt "$peak" ] && peak=$s',
  '  sleep 0.1',
  'done',
  'echo "$(grep -v "^starting worker" "$out" | tail -1) $peak"',
  'rm -f "$out"'
), sampler)

measure <- function(version, what, rdata, cores, n) {
  result <- tempfile(fileext = ".rds")
  line <- system2("bash", c(sampler, job, what, shQuote(rdata), cores, result, n),
                  env = paste0("R_LIBS=", libs[[version]]), stdout = TRUE)
  fields <- strsplit(trimws(tail(line, 1)), " +")[[1]]
  list(seconds = as.numeric(fields[1]), peak_MB = as.numeric(fields[2]), result = readRDS(result))
}
```

## The stimulus files

Two synthetic files at 512 pixels and the default 5 scales: 1400 trials for the participant loop, 200 for the others.

```{r files}
n <- 512
x <- matrix(seq(-1, 1, length.out = n), n, n, byrow = TRUE)
y <- -matrix(seq(-1, 1, length.out = n), n, n)
face <- exp(-(x^2 / 0.45 + y^2 / 0.75))
base_png <- tempfile(fileext = ".png")
png::writePNG((face - min(face)) / diff(range(face)), base_png)

.libPaths(c(libs[["branch"]], .libPaths()))
Sys.setenv(R_LIBS = libs[["branch"]])
library(rcicr)
stimulus_file <- function(n_trials) {
  dir <- tempfile("stimuli")
  invisible(capture.output(generateStimuli2IFC(list(face = base_png), n_trials = n_trials, img_size = n,
                                               stimulus_path = dir, seed = 1, ncores = 4,
                                               save_as_png = FALSE)))
  list.files(dir, "\\.Rdata$", full.names = TRUE)
}
rdata_1400 <- stimulus_file(1400)
rdata_200 <- stimulus_file(200)
```

## One render, double and integer indices

```{r render}
stored <- new.env()
load(rdata_200, envir = stored)
double_basis <- stored$p
integer_basis <- rcicr:::renderingBasis(double_basis)
params <- stored$stimuli_params$face
c(double_MB = as.numeric(object.size(double_basis$patchIdx)) / 2^20,
  integer_MB = as.numeric(object.size(integer_basis$patchIdx)) / 2^20)
identical(lapply(1:20, function(i) generateNoiseImage(params[i, ], double_basis)),
          lapply(1:20, function(i) generateNoiseImage(params[i, ], integer_basis)))
renders <- sapply(1:2, function(run) {
  c(double = system.time(for (i in 1:20) generateNoiseImage(params[i, ], double_basis))[["elapsed"]],
    integer = system.time(for (i in 1:20) generateNoiseImage(params[i, ], integer_basis))[["elapsed"]])
})
renders
rm(stored, double_basis, integer_basis, params)
```

## The participant loop

`generateCI(participants = )` with 1400 trials split evenly between the participants.

```{r participants}
grid <- expand.grid(cores = c(1, 2, 4), participants = c(4, 20, 100))
participant_rows <- do.call(rbind, lapply(seq_len(nrow(grid)), function(i) {
  runs <- lapply(c("main", "branch"), measure, what = "participants", rdata = rdata_1400,
                 cores = grid$cores[i], n = grid$participants[i])
  data.frame(participants = grid$participants[i], cores = grid$cores[i],
             main_s = runs[[1]]$seconds, branch_s = runs[[2]]$seconds,
             main_MB = runs[[1]]$peak_MB, branch_MB = runs[[2]]$peak_MB,
             identical = identical(runs[[1]]$result, runs[[2]]$result))
}))
knitr::kable(participant_rows)
```

## The other loops

200 trials each: generating the stimuli, the t-test z-map, and an `"images"` reference distribution of 100 draws.

```{r loops}
loop_rows <- do.call(rbind, lapply(c("stimuli", "zmap", "reference"), function(what) {
  rdata <- if (what == "stimuli") base_png else rdata_200
  do.call(rbind, lapply(c(1, 4), function(cores) {
    runs <- lapply(c("main", "branch"), measure, what = what, rdata = rdata, cores = cores, n = 200)
    data.frame(loop = what, cores = cores,
               main_s = runs[[1]]$seconds, branch_s = runs[[2]]$seconds,
               main_MB = runs[[1]]$peak_MB, branch_MB = runs[[2]]$peak_MB,
               identical = identical(runs[[1]]$result, runs[[2]]$result))
  }))
}))
knitr::kable(loop_rows)
```

## What a worker costs

The extra peak memory of a run on 4 cores over the same run on 1 core, divided by the 3 extra processes. Each run holds the parent too, so this is the price of a worker, not of a process.

```{r per-worker}
per_worker <- function(rows, by) {
  serial <- rows[rows$cores == 1, ]
  parallel_rows <- rows[rows$cores == 4, ]
  data.frame(run = parallel_rows[[by]], main_MB = round((parallel_rows$main_MB - serial$main_MB) / 3),
             branch_MB = round((parallel_rows$branch_MB - serial$branch_MB) / 3), row.names = NULL)
}
knitr::kable(rbind(per_worker(participant_rows, "participants"), per_worker(loop_rows, "loop")))
```

## What this shows, and what it does not

These are one machine's numbers from one run each; another machine's will differ, and the ratios between `main` and the branch are what carries over. The `identical` columns are `TRUE` for every configuration, so the change moves no number in any of them.
